---
title: "Spatial Composition"
subtitle: "odin2/dust2 Vignette 27 (R)"
format:
  html:
    toc: true
    code-fold: false
  gfm: default
---

## Introduction

This is the R companion to the Julia spatial composition vignette. While R's odin2 does not have a categorical composition layer, we demonstrate equivalent 3-patch metapopulation models built manually using the odin DSL with arrays.

## Setup

```{r}
library(odin2)
library(dust2)
```

## 3-Patch SIR with Ring Migration

In Julia, this model was built by stratifying a base SIR and composing with migration sub-models. In R, we write the full model with array indexing:

```{r}
sir_spatial <- odin({
  n_patch <- parameter(3)

  dim(S) <- n_patch
  dim(I) <- n_patch
  dim(R) <- n_patch
  dim(foi) <- n_patch
  dim(S0) <- n_patch
  dim(I0) <- n_patch
  dim(contact) <- c(n_patch, n_patch)
  dim(mig) <- c(n_patch, n_patch)
  dim(weighted_inf) <- c(n_patch, n_patch)
  dim(mig_out_S) <- n_patch
  dim(mig_in_S) <- n_patch
  dim(mig_flux_S) <- c(n_patch, n_patch)

  # Force of infection per patch
  weighted_inf[, ] <- contact[i, j] * I[j]
  foi[] <- beta * sum(weighted_inf[i, ]) / N

  # Migration fluxes for susceptibles
  mig_flux_S[, ] <- mig[i, j] * S[j]
  mig_out_S[] <- sum(mig[, i]) * S[i]
  mig_in_S[] <- sum(mig_flux_S[i, ])

  deriv(S[]) <- -foi[i] * S[i] - mig_out_S[i] + mig_in_S[i]
  deriv(I[]) <- foi[i] * S[i] - gamma * I[i]
  deriv(R[]) <- gamma * I[i]

  initial(S[]) <- S0[i]
  initial(I[]) <- I0[i]
  initial(R[]) <- 0

  S0[] <- parameter()
  I0[] <- parameter()
  contact[, ] <- parameter()
  mig[, ] <- parameter()
  beta <- parameter(0.3)
  gamma <- parameter(0.1)
  N <- parameter(1000)
})
```

### Ring topology

```{r}
C_ring <- matrix(c(1.0, 0.1, 0.1,
                    0.1, 1.0, 0.1,
                    0.1, 0.1, 1.0), 3, 3, byrow = TRUE)

# Ring migration (symmetric): every patch connected to every other
M_ring <- matrix(c(0.0,  0.01, 0.01,
                    0.01, 0.0,  0.01,
                    0.01, 0.01, 0.0), 3, 3, byrow = TRUE)

pars_ring <- list(
  n_patch = 3,
  S0 = c(320, 330, 330),
  I0 = c(10, 0, 0),
  contact = C_ring,
  mig = M_ring,
  beta = 0.3, gamma = 0.1, N = 1000
)

sys <- dust_system_create(sir_spatial, pars_ring, ode_control = dust_ode_control())
dust_system_set_state_initial(sys)
times <- seq(0, 300, by = 0.5)
result_ring <- dust_system_simulate(sys, times)

cols <- c("blue", "red", "green")
patches <- c("p1", "p2", "p3")
# Infected are rows 4,5,6
plot(NULL, xlim = range(times), ylim = c(0, max(result_ring[4:6, ])),
     xlab = "Time", ylab = "Infected",
     main = "Ring Topology — Infected by Patch")
for (i in 1:3) {
  lines(times, result_ring[3 + i, ], col = cols[i], lwd = 2)
}
legend("topright", legend = patches, col = cols, lwd = 2)
```

### Star topology

```{r}
C_star <- matrix(c(1.0, 0.2, 0.2,
                    0.2, 1.0, 0.0,
                    0.2, 0.0, 1.0), 3, 3, byrow = TRUE)

# Star migration: p1↔p2 and p1↔p3 only
M_star <- matrix(c(0.0,  0.01, 0.01,
                    0.01, 0.0,  0.0,
                    0.01, 0.0,  0.0), 3, 3, byrow = TRUE)

pars_star <- list(
  n_patch = 3,
  S0 = c(320, 330, 330),
  I0 = c(10, 0, 0),
  contact = C_star,
  mig = M_star,
  beta = 0.3, gamma = 0.1, N = 1000
)

sys2 <- dust_system_create(sir_spatial, pars_star, ode_control = dust_ode_control())
dust_system_set_state_initial(sys2)
result_star <- dust_system_simulate(sys2, times)

plot(NULL, xlim = range(times), ylim = c(0, max(result_star[4:6, ])),
     xlab = "Time", ylab = "Infected",
     main = "Star Topology — Infected by Patch")
for (i in 1:3) {
  lines(times, result_star[3 + i, ], col = cols[i], lwd = 2)
}
legend("topright", legend = patches, col = cols, lwd = 2)
```

## Comparing Topologies

```{r}
par(mfrow = c(1, 2))

plot(NULL, xlim = range(times), ylim = c(0, max(result_ring[4:6, ], result_star[4:6, ])),
     xlab = "Time", ylab = "Infected", main = "Ring")
for (i in 1:3) lines(times, result_ring[3 + i, ], col = cols[i], lwd = 2)
legend("topright", legend = patches, col = cols, lwd = 2)

plot(NULL, xlim = range(times), ylim = c(0, max(result_ring[4:6, ], result_star[4:6, ])),
     xlab = "Time", ylab = "Infected", main = "Star")
for (i in 1:3) lines(times, result_star[3 + i, ], col = cols[i], lwd = 2)
legend("topright", legend = patches, col = cols, lwd = 2)

par(mfrow = c(1, 1))
```

## Summary

| Feature | Julia (Odin.jl) | R (odin2) |
|---------|-----------------|-----------|
| Base model | `sir_net()` Petri net | Manual DSL |
| Spatial structure | `stratify(sir, patches; contact=C)` | Array indexing with contact matrix |
| Migration | `compose()` with migration sub-nets | Matrix multiplication in ODEs |
| Topology change | Swap contact matrix + migration nets | Swap parameter matrices |
