---
title: "Multi-Stream Outbreak Inference"
subtitle: "odin2/dust2/monty Vignette 16 (R)"
format:
  html:
    toc: true
    code-fold: false
  gfm: default
---

## Introduction

Real-world outbreak surveillance rarely delivers a single, clean data stream.
Instead, public health agencies collect **multiple overlapping signals** — case
notifications, death reports, serological surveys, wastewater concentrations —
each with its own noise, bias and delay. Fitting to a single stream can leave
key parameters poorly identified: for instance, the reporting rate ρ and the
transmission rate β may be confounded because only the product ρ × β is
constrained by case data.

This vignette demonstrates how to:

1. Build a deterministic **SEIRD** ODE model with **two comparison operators**
2. Generate synthetic case and death data
3. Fit the model using only case data (*single-stream*)
4. Fit the model using cases **and** deaths (*multi-stream*)
5. Compare posteriors to show how additional data streams tighten inference

The approach is inspired by the
[mpoxseir](https://github.com/mrc-ide/mpoxseir) package, which fits mpox
transmission models to multiple surveillance data streams simultaneously.

```{r}
library(odin2)
library(dust2)
library(monty)
```

## Model Definition

### SEIRD dynamics

We define a deterministic SEIRD model:

- **S → E**: susceptible individuals become exposed at rate β S I / N
- **E → I**: exposed progress to infectious at rate σ (mean latent period 1/σ)
- **I → R**: infectious individuals recover at rate γ
- **I → D**: infectious individuals die at rate μ

Only a fraction ρ of new infections are **reported as cases**, and a (typically
higher) fraction ρ_d of deaths are **reported**.

At each data time point the unfilter evaluates the ODE state and compares the
instantaneous flow rates to observed counts via Poisson log-likelihoods. Each
`~` statement contributes additively to the total log-likelihood:

$$\ell = \sum_t \bigl[\log p(\text{cases}_t \mid \lambda_c(t)) + \log p(\text{deaths}_t \mid \lambda_d(t))\bigr]$$

> **Note:** Comparing the instantaneous rate against a daily count is a standard
> approximation that works well when the dynamics are smooth relative to the
> observation interval.

### Multi-stream model (cases + deaths)

```{r}
seird_multi <- odin({
  deriv(S) <- -beta * S * I / N
  deriv(E) <- beta * S * I / N - sigma * E
  deriv(I) <- sigma * E - gamma * I - mu * I
  deriv(R) <- gamma * I
  deriv(D) <- mu * I

  initial(S) <- N - E0
  initial(E) <- E0
  initial(I) <- 0
  initial(R) <- 0
  initial(D) <- 0

  # Parameters
  N <- parameter(100000)
  E0 <- parameter(10)
  beta <- parameter(0.3)
  sigma <- parameter(0.2)      # 5-day mean latent period
  gamma <- parameter(0.1)      # 10-day mean infectious period
  mu <- parameter(0.005)       # death rate from I
  rho <- parameter(0.3)        # case reporting rate
  rho_d <- parameter(0.9)      # death reporting rate

  # Two data streams
  reported_cases <- data()
  reported_deaths <- data()
  reported_cases ~ Poisson(rho * sigma * E + 1e-6)
  reported_deaths ~ Poisson(rho_d * mu * I + 1e-6)
})
```

### Single-stream model (cases only)

For comparison we define a model with **identical dynamics** but only one
comparison statement.  Because the death likelihood is absent, the data cannot
directly constrain μ.

```{r}
seird_cases <- odin({
  deriv(S) <- -beta * S * I / N
  deriv(E) <- beta * S * I / N - sigma * E
  deriv(I) <- sigma * E - gamma * I - mu * I
  deriv(R) <- gamma * I
  deriv(D) <- mu * I

  initial(S) <- N - E0
  initial(E) <- E0
  initial(I) <- 0
  initial(R) <- 0
  initial(D) <- 0

  N <- parameter(100000)
  E0 <- parameter(10)
  beta <- parameter(0.3)
  sigma <- parameter(0.2)
  gamma <- parameter(0.1)
  mu <- parameter(0.005)
  rho <- parameter(0.3)

  # Single data stream
  reported_cases <- data()
  reported_cases ~ Poisson(rho * sigma * E + 1e-6)
})
```

## Synthetic Data

### True parameters

```{r}
true_pars <- list(
  N = 100000,
  E0 = 10,
  beta = 0.3,
  sigma = 0.2,
  gamma = 0.1,
  mu = 0.005,
  rho = 0.3,
  rho_d = 0.9
)

R0 <- true_pars$beta / (true_pars$gamma + true_pars$mu)
cfr <- true_pars$mu / (true_pars$gamma + true_pars$mu)
cat("R0 =", round(R0, 2), "\n")
cat("Case fatality ratio =", round(100 * cfr, 1), "%\n")
```

### Simulate the ODE

```{r}
times <- seq(0, 120, by = 1)

sys <- dust_system_create(seird_multi, true_pars, seed = 1)
dust_system_set_state_initial(sys)
result <- dust_system_simulate(sys, times)

true_S <- result[1, ]
true_E <- result[2, ]
true_I <- result[3, ]
true_D <- result[5, ]
```

```{r}
plot(times, true_I, type = "l", lwd = 2, col = "steelblue",
     xlab = "Day", ylab = "Individuals",
     main = "True SEIRD Dynamics",
     ylim = c(0, max(true_I) * 1.1))
lines(times, true_E, lwd = 2, col = "orange")
lines(times, true_D, lwd = 2, lty = 2, col = "red")
legend("topright", c("Infectious (I)", "Exposed (E)", "Cumulative Deaths (D)"),
       col = c("steelblue", "orange", "red"), lwd = 2, lty = c(1, 1, 2))
```

### Generate observations

At each day we draw reported cases and deaths from the instantaneous
flow rates, matching the comparison model.

```{r}
set.seed(42)
obs_times <- times[-1]
E_obs <- true_E[-1]
I_obs <- true_I[-1]

expected_cases  <- true_pars$rho   * true_pars$sigma * E_obs
expected_deaths <- true_pars$rho_d * true_pars$mu    * I_obs

obs_cases  <- rpois(length(expected_cases),  pmax(expected_cases,  1e-10))
obs_deaths <- rpois(length(expected_deaths), pmax(expected_deaths, 1e-10))
```

```{r}
plot(obs_times, obs_cases, pch = 16, cex = 0.6, col = "steelblue",
     xlab = "Day", ylab = "Count",
     main = "Observed Data (Two Streams)",
     ylim = c(0, max(obs_cases) * 1.1))
points(obs_times, obs_deaths, pch = 16, cex = 0.6, col = "red")
legend("topright", c("Reported cases", "Reported deaths"),
       col = c("steelblue", "red"), pch = 16)
```

### Prepare data objects

The multi-stream data includes both columns; the single-stream data only cases.

```{r}
data_multi <- data.frame(
  time = obs_times,
  reported_cases = obs_cases,
  reported_deaths = obs_deaths
)

data_cases <- data.frame(
  time = obs_times,
  reported_cases = obs_cases
)
```

## Priors

We fit three parameters — β (transmission rate), ρ (case reporting fraction),
and μ (death rate) — keeping all others fixed at their true values.

```{r}
prior <- monty_dsl({
  beta ~ Gamma(shape = 3, rate = 10)    # mean 0.3
  rho  ~ Beta(3, 7)                     # mean 0.3
  mu   ~ Gamma(shape = 2, rate = 400)   # mean 0.005
})
```

We need separate packers because the two models have different parameter sets:

```{r}
packer_cases <- monty_packer(
  c("beta", "rho", "mu"),
  fixed = list(N = 100000, E0 = 10, sigma = 0.2, gamma = 0.1))

packer_multi <- monty_packer(
  c("beta", "rho", "mu"),
  fixed = list(N = 100000, E0 = 10, sigma = 0.2, gamma = 0.1, rho_d = 0.9))
```

## Single-Stream Inference (Cases Only)

```{r}
uf_cases <- dust_unfilter_create(seird_cases, time_start = 0, data = data_cases)
ll_cases <- dust_likelihood_monty(uf_cases, packer_cases)
posterior_cases <- ll_cases + prior

cat("Log-likelihood at true parameters (cases only):",
    monty_model_density(ll_cases, c(0.3, 0.3, 0.005)), "\n")
```

```{r}
vcv <- diag(c(0.001, 0.005, 0.000001))
sampler <- monty_sampler_adaptive(vcv)

samples_cases <- monty_sample(posterior_cases, sampler, 5000,
                              initial = c(0.3, 0.3, 0.005),
                              n_chains = 1, burnin = 1000)
```

## Multi-Stream Inference (Cases + Deaths)

```{r}
uf_multi <- dust_unfilter_create(seird_multi, time_start = 0, data = data_multi)
ll_multi <- dust_likelihood_monty(uf_multi, packer_multi)
posterior_multi <- ll_multi + prior

cat("Log-likelihood at true parameters (multi-stream):",
    monty_model_density(ll_multi, c(0.3, 0.3, 0.005)), "\n")
```

```{r}
samples_multi <- monty_sample(posterior_multi, sampler, 5000,
                              initial = c(0.3, 0.3, 0.005),
                              n_chains = 1, burnin = 1000)
```

## Posterior Comparison

```{r}
beta_c <- samples_cases$pars[1, , 1]
rho_c  <- samples_cases$pars[2, , 1]
mu_c   <- samples_cases$pars[3, , 1]

beta_m <- samples_multi$pars[1, , 1]
rho_m  <- samples_multi$pars[2, , 1]
mu_m   <- samples_multi$pars[3, , 1]
```

```{r}
summarise_param <- function(name, vals, truth) {
  m  <- signif(mean(vals), 3)
  lo <- signif(quantile(vals, 0.025), 3)
  hi <- signif(quantile(vals, 0.975), 3)
  w  <- signif(hi - lo, 3)
  cat(sprintf("  %s: %s [%s, %s]  (width=%s, true=%s)\n",
              name, m, lo, hi, w, truth))
}

cat("Single-stream (cases only):\n")
summarise_param("beta", beta_c, 0.3)
summarise_param("rho",  rho_c,  0.3)
summarise_param("mu",   mu_c,   0.005)

cat("\nMulti-stream (cases + deaths):\n")
summarise_param("beta", beta_m, 0.3)
summarise_param("rho",  rho_m,  0.3)
summarise_param("mu",   mu_m,   0.005)
```

### Marginal posterior densities

```{r}
par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

# beta
hist(beta_c, breaks = 40, col = adjustcolor("steelblue", 0.5),
     probability = TRUE, main = expression("Posterior: " * beta * " (transmission)"),
     xlab = expression(beta), xlim = range(c(beta_c, beta_m)))
hist(beta_m, breaks = 40, col = adjustcolor("orange", 0.5),
     probability = TRUE, add = TRUE)
abline(v = 0.3, col = "red", lwd = 2)
legend("topright", c("Cases only", "Cases + deaths", "True"),
       fill = c(adjustcolor("steelblue", 0.5),
                adjustcolor("orange", 0.5), NA),
       border = c("black", "black", NA),
       lwd = c(NA, NA, 2), col = c(NA, NA, "red"), cex = 0.8)

# rho
hist(rho_c, breaks = 40, col = adjustcolor("steelblue", 0.5),
     probability = TRUE, main = expression("Posterior: " * rho * " (reporting)"),
     xlab = expression(rho), xlim = range(c(rho_c, rho_m)))
hist(rho_m, breaks = 40, col = adjustcolor("orange", 0.5),
     probability = TRUE, add = TRUE)
abline(v = 0.3, col = "red", lwd = 2)

# mu
hist(mu_c, breaks = 40, col = adjustcolor("steelblue", 0.5),
     probability = TRUE, main = expression("Posterior: " * mu * " (death rate)"),
     xlab = expression(mu), xlim = range(c(mu_c, mu_m)))
hist(mu_m, breaks = 40, col = adjustcolor("orange", 0.5),
     probability = TRUE, add = TRUE)
abline(v = 0.005, col = "red", lwd = 2)
```

The death data is particularly informative for μ and ρ:

- **μ (death rate)**: With cases only, μ is weakly identified because deaths
  are not directly observed.  Adding the death stream pins μ down.
- **ρ (reporting rate)**: Case counts constrain the product ρ × β but not ρ
  individually.  Deaths provide an independent signal about the absolute
  infection level, breaking this confounding.
- **β (transmission rate)**: Also tightens because ρ and β are less correlated
  once both streams contribute.

### Joint β–ρ posterior

```{r}
par(mfrow = c(1, 1))
plot(beta_c, rho_c, pch = 16, cex = 0.3,
     col = adjustcolor("steelblue", 0.15),
     xlab = expression(beta), ylab = expression(rho),
     main = expression("Joint Posterior: " * beta * " vs " * rho),
     xlim = range(c(beta_c, beta_m)),
     ylim = range(c(rho_c, rho_m)))
points(beta_m, rho_m, pch = 16, cex = 0.3,
       col = adjustcolor("orange", 0.15))
points(0.3, 0.3, pch = 8, cex = 2, col = "red", lwd = 2)
legend("topright", c("Cases only", "Cases + deaths", "True"),
       col = c("steelblue", "orange", "red"),
       pch = c(16, 16, 8), cex = 0.8)
```

The single-stream posterior shows a strong **negative correlation** between β
and ρ — higher transmission with lower reporting can produce the same case
count.  The multi-stream posterior is tighter and better centred because the
death data breaks this degeneracy.

## Forecasting from the Posterior

We project the epidemic forward using posterior draws from each fit,
propagating parameter uncertainty into forecasts.

```{r}
proj_times <- seq(0, 180, by = 1)
n_proj <- 200
n_t <- length(proj_times)

project <- function(samples, n_proj, seed) {
  set.seed(seed)
  n_total <- dim(samples$pars)[2]
  idx <- sample(n_total, n_proj, replace = TRUE)

  I_traj <- matrix(NA, n_proj, n_t)
  D_traj <- matrix(NA, n_proj, n_t)

  for (j in seq_len(n_proj)) {
    pars <- list(
      N = 100000, E0 = 10, sigma = 0.2,
      gamma = 0.1, rho_d = 0.9,
      beta  = samples$pars[1, idx[j], 1],
      rho   = samples$pars[2, idx[j], 1],
      mu    = samples$pars[3, idx[j], 1]
    )
    sys <- dust_system_create(seird_multi, pars, seed = j)
    dust_system_set_state_initial(sys)
    r <- dust_system_simulate(sys, proj_times)
    I_traj[j, ] <- r[3, ]
    D_traj[j, ] <- r[5, ]
  }

  list(I = I_traj, D = D_traj)
}

proj_single <- project(samples_cases, n_proj, 10)
proj_multi  <- project(samples_multi, n_proj, 10)
```

```{r}
ribbon_plot <- function(times, traj, col, label, add = FALSE, ...) {
  med <- apply(traj, 2, median)
  lo  <- apply(traj, 2, quantile, 0.025)
  hi  <- apply(traj, 2, quantile, 0.975)
  if (!add) {
    plot(times, med, type = "n", ylim = range(c(lo, hi)), ...)
  }
  polygon(c(times, rev(times)), c(lo, rev(hi)),
          col = adjustcolor(col, 0.2), border = NA)
  lines(times, med, col = col, lwd = 2)
}

par(mfrow = c(2, 1), mar = c(4, 4, 3, 1))

ribbon_plot(proj_times, proj_single$I, "steelblue", "Cases only",
            xlab = "Day", ylab = "Infectious (I)",
            main = "Projections: Infectious")
ribbon_plot(proj_times, proj_multi$I, "orange", "Cases + deaths", add = TRUE)
lines(times, true_I, col = "red", lwd = 2, lty = 2)
legend("topright", c("Cases only", "Cases + deaths", "Truth"),
       col = c("steelblue", "orange", "red"), lwd = 2, lty = c(1, 1, 2),
       cex = 0.8)

ribbon_plot(proj_times, proj_single$D, "steelblue", "Cases only",
            xlab = "Day", ylab = "Cumulative Deaths",
            main = "Projections: Deaths")
ribbon_plot(proj_times, proj_multi$D, "orange", "Cases + deaths", add = TRUE)
lines(times, true_D, col = "red", lwd = 2, lty = 2)
legend("topleft", c("Cases only", "Cases + deaths", "Truth"),
       col = c("steelblue", "orange", "red"), lwd = 2, lty = c(1, 1, 2),
       cex = 0.8)
```

Projections from the multi-stream fit (orange) have **narrower credible
intervals** and track the true trajectory more closely, especially for
cumulative deaths.  The single-stream fit (blue) shows wider uncertainty
because the death rate μ is poorly constrained by case data alone.

## Summary

| Aspect | Single-stream | Multi-stream |
|--------|---------------|--------------|
| **Data used** | Cases only | Cases + deaths |
| **`~` operators** | 1 | 2 |
| **μ identifiability** | Weak — prior-driven | Strong — data-driven |
| **ρ–β confounding** | Yes | Reduced |
| **Projection uncertainty** | Wide | Narrow |

### Key takeaways

1. **Each `~` statement adds a log-likelihood term.** Multiple comparison
   operators let the model learn from every available data stream.
2. **Different streams constrain different parameters.** Case data pins down
   the product ρ × β; death data independently constrains μ.
3. **Multi-stream fitting reduces posterior uncertainty** and gives more
   reliable forecasts — critical for real-time outbreak response.
4. **The workflow is the same** regardless of the number of streams: declare
   `data()` variables, write `~` comparisons, and pass a data object
   containing all columns.

| Step | API |
|------|-----|
| Define model | `odin({ … })` with `~` for each data stream |
| Prepare data | `data.frame(time = …, col1 = …, col2 = …)` |
| Likelihood | `dust_unfilter_create()` → `dust_likelihood_monty()` |
| Prior | `monty_dsl({ … })` |
| Posterior | `likelihood + prior` |
| Sample | `monty_sample(posterior, sampler, n, …)` |
