---
title: "Particle Filtering in R (CPU)"
subtitle: "R companion to the GPU particle filtering vignette"
format:
  html:
    code-fold: false
    toc: true
engine: knitr
---

## Overview

The Julia vignette (vignette 31) demonstrates GPU-accelerated particle
filtering via Odin.jl's Metal/CUDA backend system. **R does not have an
equivalent GPU particle filter** in the odin2/dust2 ecosystem, but the
CPU-based `dust2` particle filter is highly optimised with C++ and OpenMP
parallelisation.

This vignette shows the equivalent R workflow for particle filtering using
`odin2` and `dust2`, which serves as the CPU baseline.

## SIR Particle Filter in R

### Define the model

```{r}
library(odin2)
library(dust2)

sir <- odin2::odin({
  update(S) <- S - n_SI
  update(I) <- I + n_SI - n_IR
  update(R) <- R + n_IR
  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0

  p_SI <- 1 - exp(-beta * I / N * dt)
  p_IR <- 1 - exp(-gamma * dt)
  n_SI <- Binomial(S, p_SI)
  n_IR <- Binomial(I, p_IR)

  cases <- data()
  cases ~ Poisson(max(I, 1))

  N <- parameter(1000)
  I0 <- parameter(10)
  beta <- parameter(0.5)
  gamma <- parameter(0.1)
})
```

### Generate synthetic data

```{r}
pars <- list(N = 1000, I0 = 10, beta = 0.5, gamma = 0.1)

sys <- dust_system_create(sir, pars, n_particles = 1, dt = 1, seed = 1L)
dust_system_set_state_initial(sys)
times <- seq(1, 100, by = 1)
result <- dust_system_simulate(sys, times)

# dust2 returns (n_state, n_particles, n_times) or (n_state, n_times)
# depending on n_particles; extract I (state 2) appropriately
if (length(dim(result)) == 3) {
  cases_raw <- result[2, 1, ]
} else {
  cases_raw <- result[2, ]
}

data_df <- data.frame(
  time = times,
  cases = pmax(1, round(cases_raw))
)
```

### Run the particle filter

```{r}
filter <- dust_filter_create(sir, data = data_df, time_start = 0,
                              n_particles = 10000, dt = 1, seed = 42L)
ll <- dust_likelihood_run(filter, pars)
cat(sprintf("Log-likelihood: %.2f\n", ll))
```

### Benchmark

```{r}
n_reps <- 20
timings <- replicate(n_reps, {
  t0 <- proc.time()["elapsed"]
  dust_likelihood_run(filter, pars)
  proc.time()["elapsed"] - t0
})
cat(sprintf("Particle filter (10,000 particles, %d reps):\n", n_reps))
cat(sprintf("  Mean: %.1f ms\n", mean(timings) * 1000))
cat(sprintf("  SD:   %.1f ms\n", sd(timings) * 1000))
```

### Integration with monty MCMC

```{r}
library(monty)

packer <- monty_packer(c("beta", "gamma"), fixed = list(I0 = 10, N = 1000))
ll_model <- dust_likelihood_monty(filter, packer)

prior <- monty_dsl({
  beta ~ Exponential(mean = 0.5)
  gamma ~ Exponential(mean = 0.2)
})

posterior <- ll_model + prior

sampler <- monty_sampler_random_walk(vcv = diag(c(0.01, 0.001)))
```

Running the full MCMC is commented out for vignette build speed:

```r
samples <- monty_sample(posterior, sampler, n_steps = 1000, n_chains = 4)
```

## GPU acceleration: R vs Julia

| Feature | R (dust2) | Julia (Odin.jl) |
|---------|-----------|-----------------|
| Particle filter | `dust_filter_create()` | `gpu_dust_filter_create()` |
| Backend | C++ / OpenMP (CPU) | Metal / CUDA / CPU |
| Parallelism | OpenMP threads | GPU kernels + CPU resampling |
| MCMC bridge | `dust_likelihood_monty()` | `gpu_dust_likelihood_monty()` |

### Why no GPU in R?

R's C++ compilation model and memory management make GPU kernel integration
challenging. The dust2 C++ code uses OpenMP for CPU-level parallelism across
particles, which provides significant speedup on multi-core machines without
requiring GPU hardware.

For GPU-accelerated particle filtering, use the Julia implementation via
`Odin.jl`, which provides Metal (Apple Silicon) and CUDA (NVIDIA) backends.

## Summary

- R's `dust2` particle filter is well-optimised with C++ and OpenMP
- For most workloads, the CPU filter in R is sufficient
- GPU acceleration is available through the Julia implementation (`Odin.jl`)
- The R and Julia APIs follow the same conceptual design: create → run → bridge to MCMC
