---
title: "Multi-Pathogen Composition"
subtitle: "odin2/dust2 Vignette 29 (R)"
format:
  html:
    toc: true
    code-fold: false
  gfm: default
---

## Introduction

This is the R companion to the Julia multi-pathogen vignette. We build a two-pathogen co-circulation model (influenza SIR + chlamydia SIS) using odin2's DSL, including cross-immunity effects.

## Setup

```{r}
library(odin2)
library(dust2)
```

## Individual Pathogen Models

### Influenza (SIR)

```{r}
flu <- odin({
  deriv(S) <- -beta * S * I / N
  deriv(I) <- beta * S * I / N - gamma * I
  deriv(R) <- gamma * I
  initial(S) <- 900
  initial(I) <- 10
  initial(R) <- 0
  beta <- parameter(0.4)
  gamma <- parameter(0.15)
  N <- parameter(1000)
})

sys <- dust_system_create(flu, list(beta = 0.4, gamma = 0.15, N = 1000),
                          ode_control = dust_ode_control())
dust_system_set_state_initial(sys)
times <- seq(0, 200, by = 0.5)
r_flu <- dust_system_simulate(sys, times)
```

### Chlamydia (SIS)

```{r}
chl <- odin({
  deriv(S) <- -beta * S * I / N + gamma * I
  deriv(I) <- beta * S * I / N - gamma * I
  initial(S) <- 900
  initial(I) <- 10
  beta <- parameter(0.3)
  gamma <- parameter(0.05)
  N <- parameter(1000)
})

sys <- dust_system_create(chl, list(beta = 0.3, gamma = 0.05, N = 1000),
                          ode_control = dust_ode_control())
dust_system_set_state_initial(sys)
r_chl <- dust_system_simulate(sys, times)
```

```{r}
par(mfrow = c(1, 2))
plot(times, r_flu[1, ], type = "l", col = "blue", lwd = 2,
     ylim = c(0, 1000), xlab = "Time", ylab = "Population",
     main = "Flu (SIR)")
lines(times, r_flu[2, ], col = "red", lwd = 2)
lines(times, r_flu[3, ], col = "green", lwd = 2)
legend("right", c("S", "I", "R"), col = c("blue", "red", "green"), lwd = 2)

plot(times, r_chl[1, ], type = "l", col = "blue", lwd = 2,
     ylim = c(0, 1000), xlab = "Time", ylab = "Population",
     main = "Chlamydia (SIS)")
lines(times, r_chl[2, ], col = "red", lwd = 2)
legend("right", c("S", "I"), col = c("blue", "red"), lwd = 2)
par(mfrow = c(1, 1))
```

## Two-Pathogen Co-Circulation

Joint state model tracking all combinations of flu × chlamydia infection status:

```{r}
two_pathogen <- odin({
  # SS: susceptible to both
  # IS: flu-infected, chl-susceptible
  # SI: flu-susceptible, chl-infected
  # II: co-infected
  # RS: flu-recovered, chl-susceptible
  # RI: flu-recovered, chl-infected

  total_I_flu <- IS + II
  total_I_chl <- SI + II + RI

  foi_flu <- beta_flu * total_I_flu / N
  foi_chl <- beta_chl * total_I_chl / N

  deriv(SS) <- -foi_flu * SS - foi_chl * SS + gamma_chl * SI
  deriv(IS) <- foi_flu * SS - gamma_flu * IS - foi_chl * IS + gamma_chl * II
  deriv(SI) <- foi_chl * SS - foi_flu * SI - gamma_chl * SI
  deriv(II) <- foi_flu * SI + foi_chl * IS - gamma_flu * II - gamma_chl * II
  deriv(RS) <- gamma_flu * IS - foi_chl * RS + gamma_chl * RI
  deriv(RI) <- gamma_flu * II + foi_chl * RS - gamma_chl * RI

  initial(SS) <- 880
  initial(IS) <- 5
  initial(SI) <- 5
  initial(II) <- 0
  initial(RS) <- 0
  initial(RI) <- 0

  beta_flu <- parameter(0.4)
  gamma_flu <- parameter(0.15)
  beta_chl <- parameter(0.3)
  gamma_chl <- parameter(0.05)
  N <- parameter(1000)
})

pars <- list(beta_flu = 0.4, gamma_flu = 0.15,
             beta_chl = 0.3, gamma_chl = 0.05, N = 1000)
sys <- dust_system_create(two_pathogen, pars, ode_control = dust_ode_control())
dust_system_set_state_initial(sys)
times <- seq(0, 300, by = 0.5)
result <- dust_system_simulate(sys, times)
```

### Marginal Prevalence

```{r}
flu_infected <- result[2, ] + result[4, ]
chl_infected <- result[3, ] + result[4, ] + result[6, ]

plot(times, flu_infected, type = "l", col = "red", lwd = 2,
     xlab = "Time", ylab = "Infected",
     main = "Co-Circulation: Marginal Prevalence",
     ylim = c(0, max(flu_infected, chl_infected)))
lines(times, chl_infected, col = "blue", lwd = 2)
legend("topright", c("Flu (SIR)", "Chlamydia (SIS)"),
       col = c("red", "blue"), lwd = 2)
```

## Cross-Immunity

```{r}
two_pathogen_xi <- odin({
  total_I_flu <- IS + II
  total_I_chl <- SI + II + RI

  foi_flu <- beta_flu * total_I_flu / N
  foi_chl <- beta_chl * total_I_chl / N

  deriv(SS) <- -foi_flu * SS - foi_chl * SS + gamma_chl * SI
  deriv(IS) <- foi_flu * SS - gamma_flu * IS - foi_chl * IS + gamma_chl * II
  deriv(SI) <- foi_chl * SS - foi_flu * SI - gamma_chl * SI
  deriv(II) <- foi_flu * SI + foi_chl * IS - gamma_flu * II - gamma_chl * II
  deriv(RS) <- gamma_flu * IS - sigma * foi_chl * RS + gamma_chl * RI
  deriv(RI) <- gamma_flu * II + sigma * foi_chl * RS - gamma_chl * RI

  initial(SS) <- 880
  initial(IS) <- 5
  initial(SI) <- 5
  initial(II) <- 0
  initial(RS) <- 0
  initial(RI) <- 0

  beta_flu <- parameter(0.4)
  gamma_flu <- parameter(0.15)
  beta_chl <- parameter(0.3)
  gamma_chl <- parameter(0.05)
  sigma <- parameter(0.5)
  N <- parameter(1000)
})

sigmas <- c(1.0, 0.5, 0.1)
labels <- c("sigma=1 (none)", "sigma=0.5 (partial)", "sigma=0.1 (strong)")
cols <- c("red", "orange", "green")

plot(NULL, xlim = range(times), ylim = c(0, 200),
     xlab = "Time", ylab = "Chlamydia infected",
     main = "Chlamydia Prevalence under Cross-Immunity")

for (k in seq_along(sigmas)) {
  pars_xi <- list(beta_flu = 0.4, gamma_flu = 0.15,
                  beta_chl = 0.3, gamma_chl = 0.05,
                  sigma = sigmas[k], N = 1000)
  sys_xi <- dust_system_create(two_pathogen_xi, pars_xi,
                               ode_control = dust_ode_control())
  dust_system_set_state_initial(sys_xi)
  r_xi <- dust_system_simulate(sys_xi, times)
  chl_i <- r_xi[3, ] + r_xi[4, ] + r_xi[6, ]
  lines(times, chl_i, col = cols[k], lwd = 2)
}
legend("topright", labels, col = cols, lwd = 2)
```

## Summary

| Feature | Julia (Odin.jl) | R (odin2) |
|---------|-----------------|-----------|
| Independent pathogens | `compose(flu, chl)` | Separate models |
| Shared-population model | Manual `@odin` (joint states) | Manual DSL (joint states) |
| Cross-immunity | Parameter σ in joint model | Parameter σ in joint model |

For pathogens sharing a host population, both Julia and R require explicit construction of the joint state space. The categorical framework helps structure thinking even when `compose()` cannot directly produce the joint model.
