---
title: "SARS-CoV-2 Multi-Region: Coupled Epidemics with Shared Parameters (R)"
subtitle: "odin2/dust2/monty Vignette 19 (R)"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
---

## Introduction

In real epidemics, disease does not respect administrative boundaries.
SARS-CoV-2 swept through cities before reaching surrounding suburbs and rural
areas, with each region experiencing waves of different size and timing.
Understanding these spatial dynamics requires models that capture **region-
specific transmission** (driven by population density, behaviour, and policy)
alongside **inter-region coupling** (commuter flows, travel).

This vignette demonstrates how to:

1. Build a **multi-region SEIRD** ODE model with per-region state variables
2. Use **per-region time-varying Rt** via separate interpolated trajectories
3. Include weak **inter-region coupling** through pairwise coupling parameters
4. Simulate heterogeneous epidemic waves across three regions
5. Generate synthetic weekly case data and fit the model using an **unfilter**
6. Recover region-specific Rt trajectories and shared reporting parameters via
   adaptive MCMC

The approach is inspired by the
[sarscov2-multiregion](https://github.com/mrc-ide/sarscov2-multiregion)
package, which fits SARS-CoV-2 transmission models to multiple English regions
simultaneously. Here we build a simplified but instructive version with three
regions.

```{r}
library(odin2)
library(dust2)
library(monty)
```

## Regional Setup

We model three regions with different population sizes and epidemic
trajectories:

| Region | Type | Population | Epidemic character |
|--------|------|------------|-------------------|
| 1 | Urban | 5,000,000 | Early, large wave — Rt peaks at 2.5 then drops sharply with lockdown |
| 2 | Suburban | 3,000,000 | Delayed wave — Rt rises more slowly, peaks at 2.0 |
| 3 | Rural | 1,000,000 | Small wave — Rt stays low, peaks at 1.5 |

### Rt trajectories

Each region has a piecewise-linear reproduction number that reflects different
intervention timelines:

```{r}
n_regions <- 3
Rt_times <- c(0, 30, 60, 90, 120, 180, 250, 365)
n_Rt <- length(Rt_times)

# Rt values per region at each time point
Rt_region1 <- c(2.5, 2.5, 1.2, 0.8, 1.1, 1.3, 1.5, 1.2)  # Urban: early lockdown
Rt_region2 <- c(1.0, 1.8, 2.0, 1.0, 0.9, 1.2, 1.4, 1.1)  # Suburban: delayed
Rt_region3 <- c(1.0, 1.2, 1.5, 1.3, 0.9, 1.0, 1.1, 1.0)  # Rural: mild

plot(Rt_times, Rt_region1, type = "b", pch = 16, col = "steelblue", lwd = 2,
     xlab = "Day", ylab = "Rt", main = "Prescribed Rt Trajectories",
     ylim = c(0.5, 3.0))
lines(Rt_times, Rt_region2, type = "b", pch = 15, col = "orange", lwd = 2)
lines(Rt_times, Rt_region3, type = "b", pch = 17, col = "forestgreen", lwd = 2)
abline(h = 1, lty = 2, col = "gray")
legend("topright",
       legend = c("Region 1 (Urban)", "Region 2 (Suburban)", "Region 3 (Rural)"),
       col = c("steelblue", "orange", "forestgreen"), pch = c(16, 15, 17), lwd = 2)
```

### Inter-region coupling

Coupling between regions is captured by pairwise parameters.  The coupling
strengths are small, representing commuter or travel-based importation:

- Urban ↔ Suburban: `c12 = 0.1` (strongest, commuter flows)
- Urban ↔ Rural: `c13 = 0.05` (weaker)
- Suburban ↔ Rural: `c23 = 0.1`

A global scaling parameter `epsilon = 0.05` controls the overall importation
pressure.

## Model Definition

Since we have three regions with distinct Rt trajectories, we define a separate
interpolation for each region's Rt.  With only three regions, we expand the
SEIRD dynamics explicitly per region rather than using arrays.

The **force of infection** in region *i* combines local transmission with a
small contribution from infectious individuals in connected regions:

$$
\lambda_i = \beta_i \left( \frac{I_i}{N_i} + \varepsilon \sum_j C_{ij} \frac{I_j}{N_j} \right)
$$

where $\beta_i = \text{Rt}_i \times \gamma$, $\varepsilon$ is the coupling
strength, and $C_{ij}$ are the pairwise coupling coefficients.

```{r}
seird_regions <- odin({
  # Per-region Rt via interpolation
  n_Rt_times <- parameter()
  dim(Rt_t) <- n_Rt_times
  dim(Rt_v1) <- n_Rt_times
  dim(Rt_v2) <- n_Rt_times
  dim(Rt_v3) <- n_Rt_times
  Rt_t[] <- parameter()
  Rt_v1[] <- parameter()
  Rt_v2[] <- parameter()
  Rt_v3[] <- parameter()
  Rt_1 <- interpolate(Rt_t, Rt_v1, "linear")
  Rt_2 <- interpolate(Rt_t, Rt_v2, "linear")
  Rt_3 <- interpolate(Rt_t, Rt_v3, "linear")

  # Epidemiological parameters
  gamma <- parameter(0.2)
  sigma <- parameter(0.333)
  ifr <- parameter(0.005)
  rho <- parameter(0.3)
  epsilon <- parameter(0.05)

  # Population and initial conditions
  N1 <- parameter(5e6)
  N2 <- parameter(3e6)
  N3 <- parameter(1e6)
  I0_1 <- parameter(100)
  I0_2 <- parameter(10)
  I0_3 <- parameter(1)

  # Coupling (symmetric pairwise)
  c12 <- parameter(0.1)
  c13 <- parameter(0.05)
  c23 <- parameter(0.1)

  # Force of infection per region
  foi_1 <- Rt_1 * gamma * (I1 / N1 + epsilon * (c12 * I2 / N2 + c13 * I3 / N3))
  foi_2 <- Rt_2 * gamma * (I2 / N2 + epsilon * (c12 * I1 / N1 + c23 * I3 / N3))
  foi_3 <- Rt_3 * gamma * (I3 / N3 + epsilon * (c13 * I1 / N1 + c23 * I2 / N2))

  # Region 1 (Urban)
  deriv(S1) <- -foi_1 * S1
  deriv(E1) <- foi_1 * S1 - sigma * E1
  deriv(I1) <- sigma * E1 - gamma * I1
  deriv(R1) <- (1 - ifr) * gamma * I1
  deriv(D1) <- ifr * gamma * I1
  deriv(cum1) <- foi_1 * S1

  # Region 2 (Suburban)
  deriv(S2) <- -foi_2 * S2
  deriv(E2) <- foi_2 * S2 - sigma * E2
  deriv(I2) <- sigma * E2 - gamma * I2
  deriv(R2) <- (1 - ifr) * gamma * I2
  deriv(D2) <- ifr * gamma * I2
  deriv(cum2) <- foi_2 * S2

  # Region 3 (Rural)
  deriv(S3) <- -foi_3 * S3
  deriv(E3) <- foi_3 * S3 - sigma * E3
  deriv(I3) <- sigma * E3 - gamma * I3
  deriv(R3) <- (1 - ifr) * gamma * I3
  deriv(D3) <- ifr * gamma * I3
  deriv(cum3) <- foi_3 * S3

  # Initial conditions
  initial(S1) <- N1 - I0_1
  initial(E1) <- 0
  initial(I1) <- I0_1
  initial(R1) <- 0
  initial(D1) <- 0
  initial(cum1) <- 0

  initial(S2) <- N2 - I0_2
  initial(E2) <- 0
  initial(I2) <- I0_2
  initial(R2) <- 0
  initial(D2) <- 0
  initial(cum2) <- 0

  initial(S3) <- N3 - I0_3
  initial(E3) <- 0
  initial(I3) <- I0_3
  initial(R3) <- 0
  initial(D3) <- 0
  initial(cum3) <- 0

  # Outputs
  output(daily_cases_1) <- rho * sigma * E1
  output(daily_cases_2) <- rho * sigma * E2
  output(daily_cases_3) <- rho * sigma * E3
  output(sero_1) <- cum1 / N1
  output(sero_2) <- cum2 / N2
  output(sero_3) <- cum3 / N3

  # Data comparison
  cases1 <- data()
  cases2 <- data()
  cases3 <- data()
  cases1 ~ Poisson(rho * sigma * E1 * 7 + 1e-6)
  cases2 ~ Poisson(rho * sigma * E2 * 7 + 1e-6)
  cases3 ~ Poisson(rho * sigma * E3 * 7 + 1e-6)
})
```

**Key features of this model:**

- **Explicit per-region states**: `S1`, `E1`, `I1`, etc. are written out for
  each of the three regions, avoiding DSL array edge cases while keeping the
  dynamics identical
- **Separate Rt interpolation**: Each region has its own piecewise-linear Rt
  trajectory, reflecting different intervention timings
- **Pairwise coupling**: The `c12`, `c13`, `c23` parameters and `epsilon`
  control importation pressure between regions
- **Two outputs per region**: `daily_cases_*` (reported case rate) and
  `sero_*` (cumulative attack rate) for monitoring

## Simulation

### Parameter setup

```{r}
true_pars <- list(
  n_Rt_times = n_Rt,
  Rt_t = Rt_times,
  Rt_v1 = Rt_region1,
  Rt_v2 = Rt_region2,
  Rt_v3 = Rt_region3,
  epsilon = 0.05,
  N1 = 5e6,
  N2 = 3e6,
  N3 = 1e6,
  I0_1 = 100,
  I0_2 = 10,
  I0_3 = 1,
  c12 = 0.1,
  c13 = 0.05,
  c23 = 0.1,
  gamma = 0.2,
  sigma = 1 / 3,
  ifr = 0.005,
  rho = 0.3
)

cat("Infectious period:", 1 / true_pars$gamma, "days\n")
cat("Latent period:", round(1 / true_pars$sigma, 1), "days\n")
cat("IFR:", 100 * true_pars$ifr, "%\n")
cat("Reporting rate:", 100 * true_pars$rho, "%\n")
```

### Run simulation

```{r}
sim_times <- seq(0, 365, by = 1)

sys <- dust_system_create(seird_regions, true_pars,
                          ode_control = dust_ode_control())
dust_system_set_state_initial(sys)
result <- dust_system_simulate(sys, sim_times)
cat("Output dimensions:", dim(result), "\n")
```

The state vector is ordered as: `S1, E1, I1, R1, D1, cum1, S2, E2, I2, R2,
D2, cum2, S3, E3, I3, R3, D3, cum3` (18 states), followed by the outputs
`daily_cases_1, daily_cases_2, daily_cases_3, sero_1, sero_2, sero_3`
(6 outputs).

### Epidemic dynamics

```{r}
region_names <- c("Urban (5M)", "Suburban (3M)", "Rural (1M)")
region_cols <- c("steelblue", "orange", "forestgreen")

# State indices: I1=3, I2=9, I3=15
I_idx <- c(3, 9, 15)

# Infectious individuals by region
plot(NULL, xlim = range(sim_times),
     ylim = c(0, max(result[I_idx, ])),
     xlab = "Day", ylab = "Infectious (I)",
     main = "Infectious Individuals by Region")
for (i in 1:n_regions) {
  lines(sim_times, result[I_idx[i], ], col = region_cols[i], lwd = 2)
}
legend("topright", region_names, col = region_cols, lwd = 2)
```

```{r}
# State indices: D1=5, D2=11, D3=17
D_idx <- c(5, 11, 17)

# Cumulative deaths by region
plot(NULL, xlim = range(sim_times),
     ylim = c(0, max(result[D_idx, ])),
     xlab = "Day", ylab = "Cumulative Deaths",
     main = "Cumulative Deaths by Region")
for (i in 1:n_regions) {
  lines(sim_times, result[D_idx[i], ], col = region_cols[i], lwd = 2)
}
legend("topleft", region_names, col = region_cols, lwd = 2)
```

```{r}
# Output indices: daily_cases_1=19, daily_cases_2=20, daily_cases_3=21
cases_idx <- c(19, 20, 21)

# Daily reported cases (output)
plot(NULL, xlim = range(sim_times),
     ylim = c(0, max(result[cases_idx, ])),
     xlab = "Day", ylab = "Daily Reported Cases",
     main = "Expected Daily Cases by Region")
for (i in 1:n_regions) {
  lines(sim_times, result[cases_idx[i], ], col = region_cols[i], lwd = 2)
}
legend("topright", region_names, col = region_cols, lwd = 2)
```

```{r}
# Output indices: sero_1=22, sero_2=23, sero_3=24
sero_idx <- c(22, 23, 24)

# Seroprevalence over time
plot(NULL, xlim = range(sim_times), ylim = c(0, 1),
     xlab = "Day", ylab = "Seroprevalence",
     main = "Cumulative Seroprevalence by Region")
for (i in 1:n_regions) {
  lines(sim_times, result[sero_idx[i], ], col = region_cols[i], lwd = 2)
}
legend("topleft", region_names, col = region_cols, lwd = 2)
```

The urban region (blue) experiences the earliest and largest wave due to its
high initial Rt of 2.5.  The suburban region (orange) follows with a delayed
peak.  The rural region (green) has a much smaller epidemic, partly because its
Rt stays closer to 1, but inter-region coupling ensures it still experiences
some transmission even when local Rt is below 1.

## Synthetic Data

We generate weekly reported case counts for each region, matching the Poisson
comparison in the model.

```{r}
set.seed(42)

# Weekly observation times
obs_weeks <- seq(7, 364, by = 7)
n_weeks <- length(obs_weeks)

# Simulate to get state at weekly intervals
sys_w <- dust_system_create(seird_regions, true_pars,
                            ode_control = dust_ode_control())
dust_system_set_state_initial(sys_w)
result_weekly <- dust_system_simulate(sys_w, obs_weeks)

# State indices for E (exposed): E1=2, E2=8, E3=14
E_idx <- c(2, 8, 14)

# Generate Poisson-distributed weekly case counts
obs_cases <- matrix(0L, n_weeks, n_regions)
for (w in seq_len(n_weeks)) {
  for (i in 1:n_regions) {
    E_i <- result_weekly[E_idx[i], w]
    expected <- true_pars$rho * true_pars$sigma * E_i * 7
    obs_cases[w, i] <- rpois(1, max(expected, 1e-10))
  }
}

cat("Total observed cases:", sum(obs_cases), "\n")
cat("Cases per region:", colSums(obs_cases), "\n")
```

```{r}
plot(NULL, xlim = range(obs_weeks),
     ylim = c(0, max(obs_cases) * 1.1),
     xlab = "Day", ylab = "Weekly Reported Cases",
     main = "Synthetic Weekly Case Data")
for (i in 1:n_regions) {
  points(obs_weeks, obs_cases[, i], pch = 16, cex = 0.7,
         col = adjustcolor(region_cols[i], 0.7))
  lines(obs_weeks, obs_cases[, i],
        col = adjustcolor(region_cols[i], 0.3))
}
legend("topright", region_names, col = region_cols, pch = 16)
```

### Prepare data for the unfilter

```{r}
data <- data.frame(
  time = obs_weeks,
  cases1 = obs_cases[, 1],
  cases2 = obs_cases[, 2],
  cases3 = obs_cases[, 3]
)
```

## Inference

### Parameters to fit

We fit the **Rt trajectory values** for each region plus the shared **reporting
rate** and **IFR**, while fixing the epidemiological time scales and coupling
structure.

| Parameter | Fitted? | Notes |
|-----------|---------|-------|
| `Rt_v1[1:8]` | ✓ | Region 1 Rt at 8 time points |
| `Rt_v2[1:8]` | ✓ | Region 2 Rt at 8 time points |
| `Rt_v3[1:8]` | ✓ | Region 3 Rt at 8 time points |
| `rho` | ✓ | Shared reporting rate |
| `ifr` | ✓ | Shared infection fatality rate |
| `gamma`, `sigma` | ✗ | Fixed at known values |
| `epsilon`, `c12`, `c13`, `c23` | ✗ | Fixed — coupling structure assumed known |
| `N1`, `N2`, `N3`, `I0_*` | ✗ | Fixed population sizes and seeds |

This gives **26 parameters** total: 8 × 3 = 24 Rt values + `rho` + `ifr`.

### Packer and fixed parameters

```{r}
packer <- monty_packer(
  c("rho", "ifr"),
  array = list(Rt_v1 = n_Rt, Rt_v2 = n_Rt, Rt_v3 = n_Rt),
  fixed = list(
    n_Rt_times = n_Rt,
    Rt_t = Rt_times,
    epsilon = 0.05,
    N1 = 5e6,
    N2 = 3e6,
    N3 = 1e6,
    I0_1 = 100,
    I0_2 = 10,
    I0_3 = 1,
    c12 = 0.1,
    c13 = 0.05,
    c23 = 0.1,
    gamma = 0.2,
    sigma = 1 / 3
  )
)

cat("Number of fitted parameters:", packer$len, "\n")
```

### Prior distributions

We place weakly informative priors on each parameter:

```{r}
# R monty_dsl does not support array priors directly.
# We build a custom prior density function instead.
prior_density <- function(theta) {
  # theta layout: rho, ifr, Rt_v1[1:8], Rt_v2[1:8], Rt_v3[1:8]
  ll <- dbeta(theta[1], 3, 7, log = TRUE) +    # rho
        dbeta(theta[2], 2, 398, log = TRUE)     # ifr
  # Normal(1.5, 0.7) priors on all 24 Rt knots
  ll <- ll + sum(dnorm(theta[3:26], 1.5, 0.7, log = TRUE))
  ll
}

prior <- monty_model(
  list(
    density = prior_density,
    parameters = packer$names(),
    domain = cbind(
      c(0, 0, rep(-Inf, 24)),   # lower
      c(1, 1, rep(Inf, 24))     # upper
    )
  )
)
```

### Likelihood via unfilter

The unfilter evaluates the ODE at each data time point and computes the
Poisson log-likelihood of the weekly case counts:

```{r}
uf <- dust_unfilter_create(seird_regions, time_start = 0, data = data)
ll <- dust_likelihood_monty(uf, packer)
posterior <- ll + prior

# Check log-likelihood at true values
true_theta <- c(true_pars$rho, true_pars$ifr,
                Rt_region1, Rt_region2, Rt_region3)
ll_true <- monty_model_density(ll, true_theta)
cat("Log-likelihood at true parameters:", round(ll_true, 2), "\n")
```

### Run MCMC

We use an adaptive random walk sampler with a diagonal initial proposal.  The
adaptation phase adjusts the proposal covariance to achieve efficient mixing
across all 26 parameters.

```{r}
n_pars <- packer$len
vcv_init <- diag(c(
  0.01, 0.000001,                     # rho, ifr
  rep(0.01, n_Rt),                    # Rt_v1
  rep(0.01, n_Rt),                    # Rt_v2
  rep(0.01, n_Rt)                     # Rt_v3
))

sampler <- monty_sampler_adaptive(vcv_init)

samples <- monty_sample(posterior, sampler, 3000,
                        initial = true_theta,
                        n_chains = 1, burnin = 500)
```

## Posterior Analysis

### Recovered Rt trajectories

The main output of interest is the posterior distribution of Rt over time for
each region.  We extract the Rt values at each time point and visualise the
posterior uncertainty:

```{r}
# Extract posterior samples (packer ordering: rho, ifr, Rt_v1[1:8], Rt_v2[1:8], Rt_v3[1:8])
rho_post <- samples$pars[1, , 1]
ifr_post <- samples$pars[2, , 1]
Rt1_post <- samples$pars[3:10, , 1]    # 8 × n_samples
Rt2_post <- samples$pars[11:18, , 1]
Rt3_post <- samples$pars[19:26, , 1]
```

```{r}
rt_ribbon <- function(times, Rt_post, true_vals, col, add = FALSE, ...) {
  med <- apply(Rt_post, 1, median)
  lo  <- apply(Rt_post, 1, quantile, 0.025)
  hi  <- apply(Rt_post, 1, quantile, 0.975)
  if (!add) {
    plot(times, med, type = "n",
         ylim = c(0, max(3.5, max(hi))), ...)
  }
  polygon(c(times, rev(times)), c(lo, rev(hi)),
          col = adjustcolor(col, 0.2), border = NA)
  lines(times, med, col = col, lwd = 2)
  points(times, true_vals, pch = 16, col = col, cex = 1.2)
}

plot(NULL, xlim = range(Rt_times), ylim = c(0, 3.5),
     xlab = "Day", ylab = "Rt",
     main = "Posterior Rt Trajectories (median + 95% CrI)")

rt_ribbon(Rt_times, Rt1_post, Rt_region1, "steelblue", add = TRUE)
rt_ribbon(Rt_times, Rt2_post, Rt_region2, "orange", add = TRUE)
rt_ribbon(Rt_times, Rt3_post, Rt_region3, "forestgreen", add = TRUE)
abline(h = 1, lty = 2, col = "gray")
legend("topright",
       legend = c("Region 1 (Urban)", "Region 2 (Suburban)", "Region 3 (Rural)"),
       col = c("steelblue", "orange", "forestgreen"), lwd = 2, pch = 16)
```

The dots show the true Rt values used to generate the data.  The credible
intervals should encompass the truth at most time points, demonstrating that
the MCMC has successfully recovered the heterogeneous regional dynamics.

### Shared parameter posteriors

```{r}
summarise_param <- function(name, vals, truth) {
  m  <- signif(mean(vals), 3)
  lo <- signif(quantile(vals, 0.025), 3)
  hi <- signif(quantile(vals, 0.975), 3)
  cat(sprintf("  %s: %s [%s, %s]  (true=%s)\n", name, m, lo, hi, truth))
}

cat("Shared parameters:\n")
summarise_param("rho (reporting)", rho_post, 0.3)
summarise_param("IFR", ifr_post, 0.005)
```

```{r}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

hist(rho_post, breaks = 40, probability = TRUE,
     main = expression("Posterior: " * rho * " (reporting)"),
     xlab = expression(rho), col = adjustcolor("steelblue", 0.5))
abline(v = 0.3, col = "red", lwd = 2)

hist(ifr_post, breaks = 40, probability = TRUE,
     main = "Posterior: IFR",
     xlab = "IFR", col = adjustcolor("steelblue", 0.5))
abline(v = 0.005, col = "red", lwd = 2)
```

### Posterior predictive check

We draw parameter sets from the posterior and simulate forward to compare the
model predictions against observed data:

```{r}
n_proj <- 100
proj_times <- seq(0, 365, by = 1)
n_t <- length(proj_times)
I_traj <- array(NA, dim = c(n_proj, n_t, n_regions))

set.seed(10)
n_total <- dim(samples$pars)[2]
idx <- sample(n_total, n_proj, replace = TRUE)

for (j in seq_len(n_proj)) {
  pars_j <- packer$unpack(samples$pars[, idx[j], 1])
  sys_j <- dust_system_create(seird_regions, pars_j,
                              ode_control = dust_ode_control())
  dust_system_set_state_initial(sys_j)
  r <- dust_system_simulate(sys_j, proj_times)
  for (i in 1:n_regions) {
    I_traj[j, , i] <- r[I_idx[i], ]
  }
}
```

```{r}
par(mfrow = c(1, 1))

plot(NULL, xlim = range(proj_times),
     ylim = c(0, max(I_traj, na.rm = TRUE) * 1.05),
     xlab = "Day", ylab = "Infectious (I)",
     main = "Posterior Predictive: Infectious by Region")

for (i in 1:n_regions) {
  med <- apply(I_traj[, , i], 2, median)
  lo  <- apply(I_traj[, , i], 2, quantile, 0.025)
  hi  <- apply(I_traj[, , i], 2, quantile, 0.975)
  polygon(c(proj_times, rev(proj_times)), c(lo, rev(hi)),
          col = adjustcolor(region_cols[i], 0.15), border = NA)
  lines(proj_times, med, col = region_cols[i], lwd = 2)
  # True trajectory
  lines(sim_times, result[I_idx[i], ], col = region_cols[i], lwd = 1.5, lty = 2)
}
legend("topright", region_names, col = region_cols, lwd = 2)
```

### Trace plots

```{r}
par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))

plot(rho_post, type = "l", col = adjustcolor("black", 0.5),
     xlab = "Iteration", ylab = expression(rho),
     main = "Trace: Reporting Rate")
abline(h = 0.3, col = "red", lwd = 2)

plot(Rt1_post[1, ], type = "l", col = adjustcolor("black", 0.5),
     xlab = "Iteration", ylab = expression(Rt[1](t == 0)),
     main = "Trace: Region 1 Rt at t=0")
abline(h = 2.5, col = "red", lwd = 2)

plot(Rt2_post[3, ], type = "l", col = adjustcolor("black", 0.5),
     xlab = "Iteration", ylab = expression(Rt[2](t == 60)),
     main = "Trace: Region 2 Rt at t=60")
abline(h = 2.0, col = "red", lwd = 2)
```

## The Role of Coupling

To see the effect of inter-region coupling, we can compare the fit with and
without the coupling term.  When coupling is present, infections "leak" between
regions, which means:

- The rural region sees cases even when its local Rt is below 1
- The timing of peaks is shifted by importation pressure
- Parameter estimates are subtly different because the coupling redistributes
  infection across regions

```{r}
par(mfrow = c(1, 1))

pars_nocoupling <- true_pars
pars_nocoupling$epsilon <- 0

sys_nc <- dust_system_create(seird_regions, pars_nocoupling,
                             ode_control = dust_ode_control())
dust_system_set_state_initial(sys_nc)
result_nc <- dust_system_simulate(sys_nc, sim_times)

plot(NULL, xlim = range(sim_times),
     ylim = c(0, max(result[I_idx, ], result_nc[I_idx, ])),
     xlab = "Day", ylab = "Infectious (I)",
     main = "Effect of Inter-Region Coupling")

for (i in 1:n_regions) {
  lines(sim_times, result[I_idx[i], ], col = region_cols[i], lwd = 2)
  lines(sim_times, result_nc[I_idx[i], ], col = region_cols[i], lwd = 1.5, lty = 2)
}
legend("topright",
       legend = c(paste(region_names, "— coupled"),
                  paste(region_names, "— isolated")),
       col = rep(region_cols, 2), lwd = rep(c(2, 1.5), each = 3),
       lty = rep(1:2, each = 3), cex = 0.8)
```

## Summary

| Aspect | Detail |
|--------|--------|
| **Model** | Multi-region SEIRD with time-varying Rt and inter-region coupling |
| **Regions** | 3 (Urban, Suburban, Rural) with different populations and Rt |
| **Time-varying Rt** | Piecewise linear via `interpolate()`, one per region |
| **Coupling** | Pairwise parameters × coupling strength ε |
| **Data** | Weekly reported cases per region (Poisson) |
| **Inference** | Deterministic ODE unfilter + adaptive MCMC |
| **Fitted parameters** | 24 Rt values + ρ + IFR = 26 total |

### Key takeaways

1. **Expanded per-region states avoid DSL array edge cases.** With a small
   fixed number of regions, writing `S1`, `E1`, `I1`, etc. explicitly keeps the
   model straightforward and avoids complex array indexing features.

2. **Separate interpolation per region** allows each area to have its own
   intervention timeline while sharing the same epidemiological parameters
   (γ, σ, IFR).

3. **Inter-region coupling** captures importation effects that pure
   region-specific models miss — especially important for regions with Rt < 1
   that still experience cases from neighbouring areas.

4. **Joint fitting across regions** enables information sharing: the reporting
   rate ρ is constrained by data from all three regions simultaneously, giving
   tighter posterior estimates than fitting each region independently.

5. **The workflow scales naturally**: the same pattern works for additional data
   streams (deaths, serology) and more complex coupling structures.

| Step | API |
|------|-----|
| Define model | `odin({ … })` with per-region states and `interpolate()` |
| Prepare data | `data.frame(time = …, cases1 = …, …)` |
| Pack parameters | `monty_packer(c("rho", "ifr"), array = list(Rt_v1 = 8, …), fixed = …)` |
| Likelihood | `dust_unfilter_create()` → `dust_likelihood_monty()` |
| Prior | `monty_dsl({ … })` |
| Posterior | `likelihood + prior` |
| Sample | `monty_sample(posterior, sampler, n, …)` |
