---
title: "Piecewise-Constant Transmission Rate (Beta Blocks)"
subtitle: "odin2/dust2/monty Vignette 22 (R)"
format:
  html:
    toc: true
    code-fold: false
  gfm: default
---

## Introduction

Many infectious disease outbreaks exhibit time-varying transmission due to
seasonality, interventions, or behavioural changes. A common approach is to
model the transmission rate β as a **piecewise-constant** (step) function
over predefined time intervals — "beta blocks".

This vignette demonstrates fitting a deterministic discrete-time SIR model
with piecewise-constant β to synthetic Poisson-observed case data using
odin2/dust2/monty.

```{r}
library(odin2)
library(dust2)
library(monty)
set.seed(42)
```

## Model Definition

The model is **deterministic** — `new_inf = S * p_SI` (continuous, not
Binomial). This makes the likelihood exact with a single particle.

```{r}
sir_beta_blocks <- odin({
  # State updates (deterministic)
  update(S) <- S - new_inf
  update(I) <- I + new_inf - new_rec
  initial(S) <- N - I0
  initial(I) <- I0

  # Incidence resets each integer time step
  initial(incidence, zero_every = 1) <- 0
  update(incidence) <- new_inf

  # Deterministic transition probabilities
  p_SI <- 1 - exp(-beta_t * I / N * dt)
  p_IR <- 1 - exp(-gamma * dt)
  new_inf <- S * p_SI
  new_rec <- I * p_IR

  # Piecewise-constant transmission rate
  beta_times[] <- parameter()
  beta_values[] <- parameter()
  dim(beta_times, beta_values) <- parameter(rank = 1)
  beta_t <- interpolate(beta_times, beta_values, "constant")

  # Fixed parameters
  N <- parameter()
  I0 <- parameter()
  gamma <- parameter()

  # Data comparison
  cases <- data()
  cases ~ Poisson(incidence + 1e-6)
})
```

## Simulation Scenario

Four seasonal β blocks over one year (365 days):

| Period | Days | β | Interpretation |
|--------|------|---|----------------|
| Winter | 0–89 | 0.4 | Moderate transmission |
| Spring | 90–179 | 0.2 | Low transmission |
| Summer | 180–269 | 0.3 | Rising transmission |
| Autumn | 270–364 | 0.5 | High transmission |

```{r}
true_pars <- list(
  beta_times = c(0, 90, 180, 270),
  beta_values = c(0.4, 0.2, 0.3, 0.5),
  N = 10000,
  I0 = 10,
  gamma = 0.1
)

times <- seq(0, 365, by = 1)

sys <- System(sir_beta_blocks, true_pars, dt = 1, seed = 1)
dust_system_set_state_initial(sys)
result <- simulate(sys, times)
```

## Visualising the Simulation

```{r}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

plot(times, result[1, ], type = "l", col = "blue", lwd = 2,
     xlab = "Day", ylab = "Population",
     main = "Deterministic SIR with Beta Blocks",
     ylim = c(0, max(result[1, ]) * 1.05))
lines(times, result[2, ], col = "red", lwd = 2)
abline(v = c(90, 180, 270), lty = 2, col = "gray")
legend("right", legend = c("S", "I"), col = c("blue", "red"), lwd = 2)

# Transmission rate profile
beta_profile <- ifelse(times < 90, 0.4,
                ifelse(times < 180, 0.2,
                ifelse(times < 270, 0.3, 0.5)))
plot(times, beta_profile, type = "s", col = "darkgreen", lwd = 2,
     xlab = "Day", ylab = expression(beta),
     main = "Transmission Rate")
```

## Generate Synthetic Case Data

```{r}
true_incidence <- result[3, -1]  # incidence at times 1..365
observed_cases <- rpois(length(true_incidence),
                        pmax(true_incidence, 1e-6))

par(mfrow = c(1, 1))
barplot(observed_cases, col = adjustcolor("steelblue", 0.6),
        border = NA, space = 0,
        xlab = "Day", ylab = "Cases",
        main = "Synthetic Epidemic Data")
lines(seq_along(true_incidence), true_incidence, col = "red", lwd = 2)
abline(v = c(90, 180, 270), lty = 2, col = "gray")
legend("topright", legend = c("Observed", "True incidence"),
       fill = c(adjustcolor("steelblue", 0.6), NA),
       border = c("steelblue", NA),
       lwd = c(NA, 2), col = c(NA, "red"))
```

## Inference Setup

### Prepare filter data

```{r}
data <- data.frame(time = times[-1], cases = observed_cases)
```

### Parameter packer

The packer maps four scalar β parameters to the `beta_values` array
expected by the model.

```{r}
packer <- Packer(
  c("beta1", "beta2", "beta3", "beta4"),
  fixed = list(
    N = 10000,
    I0 = 10,
    gamma = 0.1,
    beta_times = c(0, 90, 180, 270)
  ),
  process = function(x) {
    list(beta_values = c(x$beta1, x$beta2, x$beta3, x$beta4))
  }
)
```

### Likelihood

```{r}
filter <- Likelihood(sir_beta_blocks, time_start = 0, data = data,
                             n_particles = 1, seed = 1)
likelihood <- as_model(filter, packer)
```

### Prior

```{r}
prior <- monty_dsl({
  beta1 ~ Gamma(shape = 4, rate = 10)
  beta2 ~ Gamma(shape = 4, rate = 10)
  beta3 ~ Gamma(shape = 4, rate = 10)
  beta4 ~ Gamma(shape = 4, rate = 10)
})
```

### Posterior

```{r}
posterior <- likelihood + prior
```

### Check likelihood at truth

```{r}
theta_true <- c(0.4, 0.2, 0.3, 0.5)
cat("Log-likelihood at true parameters:",
    monty_model_density(likelihood, theta_true), "\n")
```

## Run MCMC

```{r}
vcv <- diag(rep(0.002, 4))
sampler <- random_walk(vcv)

samples <- sample(posterior, sampler, 5000,
                        initial = theta_true, n_chains = 1,
                        burnin = 500)
```

## Results

### Parameter Recovery

```{r}
par_names <- c("beta1 (winter)", "beta2 (spring)",
               "beta3 (summer)", "beta4 (autumn)")
for (i in 1:4) {
  vals <- samples$pars[i, , 1]
  cat(sprintf("  %s: %.3f [%.3f, %.3f]  (true = %.1f)\n",
              par_names[i], mean(vals),
              quantile(vals, 0.025), quantile(vals, 0.975),
              theta_true[i]))
}
```

### Trace Plots

```{r}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
for (i in 1:4) {
  plot(samples$pars[i, , 1], type = "l", col = "steelblue",
       xlab = "Iteration", ylab = par_names[i],
       main = paste("Trace:", par_names[i]))
  abline(h = theta_true[i], col = "red", lwd = 2, lty = 2)
}
```

### Posterior Distributions

```{r}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
for (i in 1:4) {
  hist(samples$pars[i, , 1], breaks = 40, probability = TRUE,
       col = adjustcolor("steelblue", 0.6), border = "white",
       main = paste("Posterior:", par_names[i]),
       xlab = par_names[i])
  abline(v = theta_true[i], col = "red", lwd = 2)
}
```

### Fitted Transmission Rate

```{r}
par(mfrow = c(1, 1))
n_draw <- 200
idx <- sample(dim(samples$pars)[2], n_draw, replace = TRUE)
plot(NULL, xlim = c(0, 365), ylim = c(0, 0.8),
     xlab = "Day", ylab = expression(beta),
     main = "Posterior Transmission Rate Profile")
for (j in 1:n_draw) {
  b <- samples$pars[, idx[j], 1]
  profile <- ifelse(times < 90, b[1],
             ifelse(times < 180, b[2],
             ifelse(times < 270, b[3], b[4])))
  lines(times, profile, col = adjustcolor("steelblue", 0.05))
}
lines(times, beta_profile, col = "red", lwd = 3, lty = 2)
legend("topright", legend = c("Posterior draws", "True"),
       col = c("steelblue", "red"), lwd = c(1, 3), lty = c(1, 2))
```

## Summary

| Feature | R syntax |
|---------|----------|
| Deterministic update | `new_inf <- S * p_SI` |
| Piecewise-constant β | `interpolate(beta_times, beta_values, "constant")` |
| Array dimensions | `dim(beta_times, beta_values) <- parameter(rank = 1)` |
| Poisson observation | `cases ~ Poisson(incidence + 1e-6)` |
| Exact likelihood | `Likelihood(..., n_particles = 1)` |
| Process function | `process = function(x) list(beta_values = c(...))` |
| Prior (monty DSL) | `beta1 ~ Gamma(shape = 4, rate = 10)` |
| MCMC | `random_walk(vcv)` |

Both the Julia and R versions produce equivalent results using the same
model structure and inference approach.
