---
title: "SIS Model with School Closure"
subtitle: "odin2/dust2/monty Vignette 21 (R)"
format:
  html:
    toc: true
    code-fold: false
  gfm: default
---

## Introduction

R companion to the Julia school closure vignette. We model an SIS epidemic
where school closures reduce the effective transmission rate via a step-function
schedule.

```{r}
library(odin2)
library(dust2)
library(monty)
```

## Model Definition

```{r}
sis <- odin({
  update(S) <- S - n_SI + n_IS
  update(I) <- I + n_SI - n_IS

  initial(S) <- N - I0
  initial(I) <- I0

  initial(incidence, zero_every = 1) <- 0
  update(incidence) <- incidence + n_SI

  schools <- interpolate(schools_time, schools_open, "constant")
  schools_time[] <- parameter()
  schools_open[] <- parameter()
  dim(schools_time, schools_open) <- parameter(rank = 1)

  beta <- ((1 - schools) * (1 - schools_modifier) + schools) * beta0
  p_SI <- 1 - exp(-beta * I / N * dt)
  p_IS <- 1 - exp(-gamma * dt)
  n_SI <- Binomial(S, p_SI)
  n_IS <- Binomial(I, p_IS)

  N <- parameter(1000)
  I0 <- parameter(10)
  beta0 <- parameter(0.2)
  gamma <- parameter(0.1)
  schools_modifier <- parameter(0.6)

  cases <- data()
  cases ~ Poisson(incidence + 1e-6)
})
```

## School Schedule and Parameters

```{r}
schools_time <- c(0, 50, 60, 120, 130, 170, 180)
schools_open <- c(1,  0,  1,   0,   1,   0,   1)

pars <- list(
  schools_time = schools_time,
  schools_open = schools_open,
  schools_modifier = 0.6,
  beta0 = 0.2,
  gamma = 0.1,
  N = 1000,
  I0 = 10
)
```

## Simulation

```{r}
times <- seq(0, 200, by = 1)

sys <- dust_system_create(sis, pars, n_particles = 20, dt = 1, seed = 42)
dust_system_set_state_initial(sys)
result <- dust_system_simulate(sys, times)

dim(result)  # n_state × n_particles × n_times
```

## Visualisation

```{r}
closure_lo <- c(50, 120, 170)
closure_hi <- c(60, 130, 180)

add_closures <- function() {
  for (i in seq_along(closure_lo)) {
    rect(closure_lo[i], par("usr")[3], closure_hi[i], par("usr")[4],
         col = rgb(1, 0.65, 0, 0.15), border = NA)
  }
}

par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))

# S
plot(NULL, xlim = c(0, 200), ylim = range(result[1, , ]),
     xlab = "Time", ylab = "Count", main = "Susceptible (S)")
add_closures()
for (i in 1:20) lines(times, result[1, i, ], col = rgb(0, 0, 1, 0.3))
mean_S <- colMeans(result[1, , ])
lines(times, mean_S, col = "blue", lwd = 2)

# I
plot(NULL, xlim = c(0, 200), ylim = range(result[2, , ]),
     xlab = "Time", ylab = "Count", main = "Infected (I)")
add_closures()
for (i in 1:20) lines(times, result[2, i, ], col = rgb(1, 0, 0, 0.3))
mean_I <- colMeans(result[2, , ])
lines(times, mean_I, col = "red", lwd = 2)

# Incidence
plot(NULL, xlim = c(0, 200), ylim = range(result[3, , ]),
     xlab = "Time", ylab = "Cases", main = "Daily Incidence")
add_closures()
for (i in 1:20) lines(times, result[3, i, ], col = rgb(0, 0.6, 0, 0.3))
mean_inc <- colMeans(result[3, , ])
lines(times, mean_inc, col = "darkgreen", lwd = 2)
```

## Comparison: With vs Without School Closure

```{r}
pars_no_closure <- modifyList(pars, list(schools_modifier = 0.0))

sys1 <- dust_system_create(sis, pars, n_particles = 20, dt = 1, seed = 1)
dust_system_set_state_initial(sys1)
r_closure <- dust_system_simulate(sys1, times)

sys2 <- dust_system_create(sis, pars_no_closure, n_particles = 20, dt = 1, seed = 1)
dust_system_set_state_initial(sys2)
r_no_closure <- dust_system_simulate(sys2, times)

par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

plot(NULL, xlim = c(0, 200), ylim = range(c(colMeans(r_closure[2, , ]),
     colMeans(r_no_closure[2, , ]))),
     xlab = "Time", ylab = "Count", main = "Infected: Closure vs No Effect")
add_closures()
lines(times, colMeans(r_closure[2, , ]), col = "red", lwd = 2)
lines(times, colMeans(r_no_closure[2, , ]), col = "gray", lwd = 2, lty = 2)
legend("topright", c("With closure (60%)", "No closure effect"),
       col = c("red", "gray"), lwd = 2, lty = 1:2)

plot(NULL, xlim = c(0, 200), ylim = range(c(colMeans(r_closure[3, , ]),
     colMeans(r_no_closure[3, , ]))),
     xlab = "Time", ylab = "Cases", main = "Incidence: Closure vs No Effect")
add_closures()
lines(times, colMeans(r_closure[3, , ]), col = "darkgreen", lwd = 2)
lines(times, colMeans(r_no_closure[3, , ]), col = "gray", lwd = 2, lty = 2)
legend("topright", c("With closure (60%)", "No closure effect"),
       col = c("darkgreen", "gray"), lwd = 2, lty = 1:2)
```

## Generate Synthetic Data

```{r}
set.seed(42)
sys_truth <- dust_system_create(sis, pars, n_particles = 1, dt = 1, seed = 101)
dust_system_set_state_initial(sys_truth)
truth <- dust_system_simulate(sys_truth, times)

true_incidence <- truth[3, -1]
observed_cases <- rpois(length(true_incidence), pmax(true_incidence, 1e-6))

par(mfrow = c(1, 1))
barplot(observed_cases, names.arg = times[-1], col = rgb(0.27, 0.51, 0.71, 0.5),
        border = NA, xlab = "Time", ylab = "Cases", main = "Synthetic Case Data")
lines(seq_along(observed_cases), true_incidence, col = "red", lwd = 2)
```

## Particle Filter

```{r}
data <- data.frame(time = times[-1], cases = observed_cases)

filter <- dust_filter_create(sis, time_start = 0, data = data,
                             n_particles = 100, seed = 42)

ll <- dust_likelihood_run(filter, pars)
cat("Log-likelihood at true parameters:", round(ll, 2), "\n")
```

### Likelihood Profile for β₀

```{r}
beta0s <- seq(0.05, 0.5, by = 0.02)
lls <- numeric(length(beta0s))
for (i in seq_along(beta0s)) {
  test_pars <- modifyList(pars, list(beta0 = beta0s[i]))
  lls[i] <- dust_likelihood_run(filter, test_pars)
}

plot(beta0s, lls, type = "b", pch = 16, cex = 0.8,
     xlab = expression(beta[0]), ylab = "Log-likelihood",
     main = expression("Likelihood Profile for " * beta[0]))
abline(v = 0.2, lty = 2, col = "red")
legend("topright", expression("True " * beta[0]), lty = 2, col = "red")
```

## Bayesian Inference

```{r}
packer <- monty_packer(c("beta0", "gamma"),
                       fixed = list(N = 1000, I0 = 10,
                                    schools_time = schools_time,
                                    schools_open = schools_open,
                                    schools_modifier = 0.6))

likelihood <- dust_likelihood_monty(filter, packer)

prior <- monty_dsl({
  beta0 ~ Gamma(shape = 2, rate = 10)
  gamma ~ Gamma(shape = 2, rate = 20)
})

posterior <- likelihood + prior
```

```{r}
vcv <- matrix(c(0.002, 0, 0, 0.001), 2, 2)
sampler <- monty_sampler_random_walk(vcv)

samples <- monty_sample(posterior, sampler, 2000, initial = c(0.15, 0.08))
```

### Posterior Summaries

```{r}
beta0_samples <- samples$pars[1, 500:2000, 1]
gamma_samples <- samples$pars[2, 500:2000, 1]

cat("β₀: mean =", round(mean(beta0_samples), 4),
    ", 95% CI = [", round(quantile(beta0_samples, 0.025), 4),
    ",", round(quantile(beta0_samples, 0.975), 4), "]\n")
cat("γ:  mean =", round(mean(gamma_samples), 4),
    ", 95% CI = [", round(quantile(gamma_samples, 0.025), 4),
    ",", round(quantile(gamma_samples, 0.975), 4), "]\n")
cat("True: β₀ = 0.2, γ = 0.1\n")
```

```{r}
par(mfrow = c(1, 2))
hist(beta0_samples, breaks = 30, main = expression("Posterior: " * beta[0]),
     xlab = expression(beta[0]), probability = TRUE)
abline(v = 0.2, col = "red", lwd = 2)

hist(gamma_samples, breaks = 30, main = expression("Posterior: " * gamma),
     xlab = expression(gamma), probability = TRUE)
abline(v = 0.1, col = "red", lwd = 2)
```

### Trace Plots

```{r}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
plot(beta0_samples, type = "l", xlab = "Iteration",
     ylab = expression(beta[0]), main = expression("Trace: " * beta[0]))
abline(h = 0.2, col = "red")

plot(gamma_samples, type = "l", xlab = "Iteration",
     ylab = expression(gamma), main = expression("Trace: " * gamma))
abline(h = 0.1, col = "red")
```

## Benchmark

```{r}
bench::mark(
  simulation = {
    s <- dust_system_create(sis, pars, n_particles = 20, dt = 1, seed = 1)
    dust_system_set_state_initial(s)
    dust_system_simulate(s, times)
  },
  filter = dust_likelihood_run(filter, pars),
  check = FALSE, min_iterations = 10
)
```
