---
title: "Age Stratification"
subtitle: "odin2/dust2 Vignette 28 (R)"
format:
  html:
    toc: true
    code-fold: false
  gfm: default
---

## Introduction

This is the R companion to the Julia age stratification vignette. We build equivalent age-stratified SIR models manually using odin2 arrays, comparing assortative, proportionate, and realistic contact patterns.

## Setup

```{r}
library(odin2)
library(dust2)
```

## Age-Stratified SIR Model

```{r}
sir_age <- odin({
  n_age <- parameter(3)
  dim(S) <- n_age
  dim(I) <- n_age
  dim(R) <- n_age
  dim(foi) <- n_age
  dim(S0) <- n_age
  dim(I0) <- n_age
  dim(contact) <- c(n_age, n_age)
  dim(weighted) <- c(n_age, n_age)

  weighted[, ] <- contact[i, j] * I[j]
  foi[] <- beta * sum(weighted[i, ]) / N

  deriv(S[]) <- -foi[i] * S[i]
  deriv(I[]) <- foi[i] * S[i] - gamma * I[i]
  deriv(R[]) <- gamma * I[i]

  initial(S[]) <- S0[i]
  initial(I[]) <- I0[i]
  initial(R[]) <- 0

  S0[] <- parameter()
  I0[] <- parameter()
  contact[, ] <- parameter()
  beta <- parameter()
  gamma <- parameter()
  N <- parameter()
})
```

## Assortative Mixing

```{r}
C_assort <- diag(3)

pars <- list(
  n_age = 3,
  S0 = c(330, 330, 330),
  I0 = c(10/3, 10/3, 10/3),
  contact = C_assort,
  beta = 0.3, gamma = 0.1, N = 1000
)

sys <- System(sir_age, pars, ode_control = dust_ode_control())
dust_system_set_state_initial(sys)
times <- seq(0, 200, by = 0.5)
r_assort <- simulate(sys, times)

cols <- c("blue", "red", "green")
groups <- c("Child", "Adult", "Elder")
plot(NULL, xlim = range(times), ylim = c(0, max(r_assort[4:6, ])),
     xlab = "Time", ylab = "Infected",
     main = "Assortative Mixing")
for (i in 1:3) lines(times, r_assort[3 + i, ], col = cols[i], lwd = 2)
legend("topright", legend = groups, col = cols, lwd = 2)
```

## Proportionate Mixing

```{r}
activity <- c(3, 2, 1)
C_prop <- outer(activity, activity)
C_prop <- C_prop / max(C_prop)

pars$contact <- C_prop
sys <- System(sir_age, pars, ode_control = dust_ode_control())
dust_system_set_state_initial(sys)
r_prop <- simulate(sys, times)

plot(NULL, xlim = range(times), ylim = c(0, max(r_prop[4:6, ])),
     xlab = "Time", ylab = "Infected",
     main = "Proportionate Mixing")
for (i in 1:3) lines(times, r_prop[3 + i, ], col = cols[i], lwd = 2)
legend("topright", legend = groups, col = cols, lwd = 2)
```

## Realistic Contact Matrix

```{r}
C_real <- matrix(c(3.0, 1.0, 0.3,
                    1.0, 2.0, 0.5,
                    0.3, 0.5, 1.5), 3, 3, byrow = TRUE)

pars$contact <- C_real
pars$beta <- 0.15
sys <- System(sir_age, pars, ode_control = dust_ode_control())
dust_system_set_state_initial(sys)
r_real <- simulate(sys, times)

plot(NULL, xlim = range(times), ylim = c(0, max(r_real[4:6, ])),
     xlab = "Time", ylab = "Infected",
     main = "Realistic Contact Matrix")
for (i in 1:3) lines(times, r_real[3 + i, ], col = cols[i], lwd = 2)
legend("topright", legend = groups, col = cols, lwd = 2)
```

## Comparison

```{r}
par(mfrow = c(1, 3))
for (label in c("Assortative", "Proportionate", "Realistic")) {
  r <- switch(label,
              Assortative = r_assort,
              Proportionate = r_prop,
              Realistic = r_real)
  plot(NULL, xlim = range(times), ylim = c(0, max(r_assort[4:6, ], r_prop[4:6, ], r_real[4:6, ])),
       xlab = "Time", ylab = "Infected", main = label)
  for (i in 1:3) lines(times, r[3 + i, ], col = cols[i], lwd = 2)
  legend("topright", legend = groups, col = cols, lwd = 2, cex = 0.8)
}
par(mfrow = c(1, 1))
```

## Targeted Vaccination (Children Only)

```{r}
sir_age_vax <- odin({
  n_age <- parameter(3)
  dim(S) <- n_age
  dim(I) <- n_age
  dim(R) <- n_age
  dim(V) <- n_age
  dim(foi) <- n_age
  dim(S0) <- n_age
  dim(I0) <- n_age
  dim(vax_rate) <- n_age
  dim(contact) <- c(n_age, n_age)
  dim(weighted) <- c(n_age, n_age)

  weighted[, ] <- contact[i, j] * I[j]
  foi[] <- beta * sum(weighted[i, ]) / N

  deriv(S[]) <- -foi[i] * S[i] - vax_rate[i] * S[i]
  deriv(I[]) <- foi[i] * S[i] - gamma * I[i]
  deriv(R[]) <- gamma * I[i]
  deriv(V[]) <- vax_rate[i] * S[i]

  initial(S[]) <- S0[i]
  initial(I[]) <- I0[i]
  initial(R[]) <- 0
  initial(V[]) <- 0

  S0[] <- parameter()
  I0[] <- parameter()
  contact[, ] <- parameter()
  vax_rate[] <- parameter()
  beta <- parameter()
  gamma <- parameter()
  N <- parameter()
})

pars_vax <- list(
  n_age = 3,
  S0 = c(330, 330, 330),
  I0 = c(10/3, 10/3, 10/3),
  contact = C_real,
  vax_rate = c(0.01, 0, 0),
  beta = 0.15, gamma = 0.1, N = 1000
)

# No vaccination baseline
pars_nv <- pars_vax
pars_nv$vax_rate <- c(0, 0, 0)

sys_nv <- System(sir_age_vax, pars_nv, ode_control = dust_ode_control())
dust_system_set_state_initial(sys_nv)
r_nv <- simulate(sys_nv, times)

sys_v <- System(sir_age_vax, pars_vax, ode_control = dust_ode_control())
dust_system_set_state_initial(sys_v)
r_v <- simulate(sys_v, times)

par(mfrow = c(1, 2))
plot(NULL, xlim = range(times), ylim = c(0, max(r_nv[4:6, ])),
     xlab = "Time", ylab = "Infected", main = "No Vaccination")
for (i in 1:3) lines(times, r_nv[3 + i, ], col = cols[i], lwd = 2)
legend("topright", legend = groups, col = cols, lwd = 2)

plot(NULL, xlim = range(times), ylim = c(0, max(r_nv[4:6, ])),
     xlab = "Time", ylab = "Infected", main = "Children Vaccinated")
for (i in 1:3) lines(times, r_v[3 + i, ], col = cols[i], lwd = 2)
legend("topright", legend = groups, col = cols, lwd = 2)
par(mfrow = c(1, 1))
```

## Summary

| Feature | Julia (Odin.jl) | R (odin2) |
|---------|-----------------|-----------|
| Stratify model | `stratify(sir, groups; contact=C)` | Manual array indexing |
| Change mixing | Swap contact matrix | Swap parameter matrix |
| Add vaccination | `compose()` with vax sub-net | Add equations manually |
| Generated transitions | Automatic cross-group | Manual weighted sums |
