---
title: "Stiff ODE Models in R"
subtitle: "Solving stiff epidemiological models with deSolve"
format:
  html:
    code-fold: false
    toc: true
engine: knitr
---

## Introduction

Many epidemiological models involve **stiff** ODE systems — processes with
widely separated timescales. In R, the `deSolve` package provides several
solvers suitable for stiff systems:

| Solver | Method | Best for |
|--------|--------|----------|
| `lsoda` | Auto-switching Adams/BDF | General purpose (detects stiffness) |
| `lsode` | BDF (Gear) | Known stiff systems |
| `radau` | Implicit Runge-Kutta | Very stiff / DAE systems |

This vignette mirrors the Julia stiff ODE vignette (vignette 30) using
R's `deSolve` package, and shows the equivalent approach with `odin2`/`dust2`.

## A stiff epidemiological model

Consider an SIR model with extreme rate separation: very fast recovery
(γ = 1000/day) coupled with a moderate transmission rate (β = 2000/day).

```{r}
library(deSolve)

stiff_sir <- function(t, state, pars) {
  with(as.list(c(state, pars)), {
    infection <- beta * S * I / N
    recovery <- gamma * I
    list(c(
      dS = -infection,
      dI = infection - recovery,
      dR = recovery
    ))
  })
}

pars <- c(beta = 2000, gamma = 1000, N = 1000)
state0 <- c(S = 999, I = 1, R = 0)
times <- seq(0, 0.05, length.out = 101)
```

### Solving with `lsoda` (auto stiffness detection)

```{r}
sol_lsoda <- ode(state0, times, stiff_sir, pars, method = "lsoda")
cat(sprintf("lsoda: S(0.05) = %.4f, I(0.05) = %.6f, R(0.05) = %.4f\n",
            tail(sol_lsoda[, "S"], 1),
            tail(sol_lsoda[, "I"], 1),
            tail(sol_lsoda[, "R"], 1)))
cat(sprintf("Conservation: max |S+I+R - 1000| = %.3e\n",
            max(abs(rowSums(sol_lsoda[, c("S", "I", "R")]) - 1000))))
```

### Solving with `radau` (implicit Runge-Kutta)

```{r}
sol_radau <- ode(state0, times, stiff_sir, pars, method = "radau",
                 atol = 1e-8, rtol = 1e-8)
cat(sprintf("radau: S(0.05) = %.4f, I(0.05) = %.6f, R(0.05) = %.4f\n",
            tail(sol_radau[, "S"], 1),
            tail(sol_radau[, "I"], 1),
            tail(sol_radau[, "R"], 1)))
```

### Comparison with explicit solver (`rk4`)

```{r}
sol_rk4 <- ode(state0, times, stiff_sir, pars, method = "rk4")
cat(sprintf("rk4:   S(0.05) = %.4f, I(0.05) = %.6f, R(0.05) = %.4f\n",
            tail(sol_rk4[, "S"], 1),
            tail(sol_rk4[, "I"], 1),
            tail(sol_rk4[, "R"], 1)))

max_diff <- max(abs(sol_lsoda[, -1] - sol_radau[, -1]))
cat(sprintf("Max difference (lsoda vs radau): %.3e\n", max_diff))
```

Both implicit solvers agree closely. The explicit `rk4` may show instability
or require much smaller step sizes for this stiff system.

## Stiff ODE with odin2/dust2

In the odin2/dust2 ecosystem, stiff ODE models can be solved using the
built-in ODE solver. The `dust2` package currently uses a Dormand-Prince
solver; for very stiff systems, the `deSolve` approach shown above may be
preferable:
```r
sir <- odin2::odin({
  deriv(S) <- -beta * S * I / N
  deriv(I) <- beta * S * I / N - gamma * I
  deriv(R) <- gamma * I
  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  beta <- parameter(2000)
  gamma <- parameter(1000)
  I0 <- parameter(1)
  N <- parameter(1000)
})
```

## Classic stiff test: Robertson chemical kinetics

```{r}
robertson <- function(t, state, pars) {
  list(c(
    dy1 = -0.04 * state[1] + 1e4 * state[2] * state[3],
    dy2 =  0.04 * state[1] - 1e4 * state[2] * state[3] - 3e7 * state[2]^2,
    dy3 =  3e7 * state[2]^2
  ))
}

state0_rob <- c(y1 = 1, y2 = 0, y3 = 0)
times_rob <- c(0, 1e-5, 1e-4, 1e-3, 1e-2, 1e-1, 1, 10, 100, 1e4, 1e5)

sol_rob <- ode(state0_rob, times_rob, robertson, NULL,
               method = "lsoda", atol = 1e-10, rtol = 1e-10)
cat(sprintf("Robertson at t=1e5:\n  y1 = %.6f\n  y2 = %.2e\n  y3 = %.6f\n  Conservation: %.10f\n",
            tail(sol_rob[, "y1"], 1),
            tail(sol_rob[, "y2"], 1),
            tail(sol_rob[, "y3"], 1),
            sum(tail(sol_rob[, -1], 1))))
```

## Summary

| Feature | Julia (Odin.jl) | R (deSolve) |
|---------|-----------------|-------------|
| Stiff solver | SDIRK4 (L-stable, order 4) | `lsoda` (auto), `radau`, `lsode` |
| Jacobian | FD, ForwardDiff, analytical | FD, analytical |
| Adaptive stepping | Yes | Yes |
| Integration with inference | `Likelihood(solver=:sdirk)` | `dust2::Likelihood()` |

- R's `deSolve` is mature and well-tested for stiff systems
- `lsoda` automatically detects stiffness and switches methods
- For odin2/dust2 models, consider `deSolve` for very stiff problems
