---
title: "Complete Fitting Workflow"
subtitle: "odin2/dust2/monty Vignette 26 (R)"
format:
  html:
    toc: true
    code-fold: false
  gfm: default
---

## Introduction

R companion to the Julia fitting workflow vignette. This demonstrates the full
Bayesian fitting pipeline using odin2, dust2, and monty:

1. **Part 1** — Deterministic SIR fitting with unfilter + MCMC
2. **Part 2** — Stochastic SIS fitting with particle filter + MCMC
3. **Part 3** — Counterfactual projections from the posterior

```{r}
library(odin2)
library(dust2)
library(monty)
```

## Part 1: SIR ODE Fitting with Unfilter

### Data

```{r}
incidence <- data.frame(
  time = 1:20,
  cases = c(12, 23, 25, 36, 30, 57, 59, 62, 47, 52,
            56, 33, 34, 19, 27, 25, 15, 20, 11, 7)
)

plot(incidence, pch = 19, col = "red", main = "SIR Incidence Data")
```

### Model Definition

```{r}
sir <- odin({
  update(S) <- S - n_SI
  update(I) <- I + n_SI - n_IR
  update(R) <- R + n_IR
  update(incidence) <- incidence + n_SI

  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(incidence, zero_every = 1) <- 0

  p_SI <- 1 - exp(-beta * I / N * dt)
  p_IR <- 1 - exp(-gamma * dt)
  n_SI <- Binomial(S, p_SI)
  n_IR <- Binomial(I, p_IR)

  N <- parameter(1000)
  I0 <- parameter(10)
  beta <- parameter(0.2)
  gamma <- parameter(0.1)

  cases <- data()
  cases ~ Poisson(incidence)
})
```

### Deterministic Likelihood (Unfilter)

```{r}
unfilter <- dust_unfilter_create(sir, data = incidence, time_start = 0,
                                  dt = 0.25)
dust_likelihood_run(unfilter, list(beta = 0.4, gamma = 0.2))
```

### MCMC Setup

```{r}
packer <- monty_packer(c("beta", "gamma"))

likelihood <- dust_likelihood_monty(unfilter, packer)

prior <- monty_dsl({
  beta ~ Exponential(mean = 0.5)
  gamma ~ Exponential(mean = 0.3)
})

posterior <- likelihood + prior
```

### Run MCMC (3 chains)

```{r}
vcv <- matrix(c(0.01, 0.005, 0.005, 0.005), 2, 2)
sampler <- monty_sampler_random_walk(vcv)

samples_det <- monty_sample(posterior, sampler, 2000, n_chains = 3)
```

### Posterior Summary

```{r}
samples_df <- posterior::as_draws_df(samples_det)
posterior::summarise_draws(samples_df)
```

### Trace Plots

```{r}
matplot(samples_det$density, type = "l", lty = 1,
        xlab = "Iteration", ylab = "Log-density",
        main = "Log-posterior trace")
```

```{r}
bayesplot::mcmc_trace(samples_df)
```

### Posterior Density

```{r}
bayesplot::mcmc_scatter(samples_df)
```

### Stochastic Comparison

```{r}
filter <- dust_filter_create(sir, data = incidence, time_start = 0,
                              n_particles = 200, dt = 0.25)

likelihood_stoch <- dust_likelihood_monty(filter, packer)
posterior_stoch <- likelihood_stoch + prior

samples_stoch <- monty_sample(posterior_stoch, sampler, 2000, n_chains = 3)

pars_det <- array(samples_det$pars, c(2, prod(dim(samples_det$pars)[-1])))
pars_stoch <- array(samples_stoch$pars, c(2, prod(dim(samples_stoch$pars)[-1])))

plot(pars_det[1, ], pars_det[2, ], xlab = "beta", ylab = "gamma",
     pch = 19, col = "red", cex = 0.3,
     main = "Stochastic vs Deterministic Posterior")
points(pars_stoch[1, ], pars_stoch[2, ], pch = 19, col = "blue", cex = 0.3)
legend("topright", c("Deterministic", "Stochastic"),
       col = c("red", "blue"), pch = 19)
```

## Part 2: SIS Stochastic Fitting with Particle Filter

### Data

```{r}
schools_data <- data.frame(
  time = 1:150,
  cases = c(2,3,2,2,4,7,2,2,0,3,1,5,4,5,4,5,14,6,12,6,
            6,9,4,7,11,19,18,25,15,16,27,15,19,27,35,23,20,32,23,32,
            30,21,58,31,40,46,38,32,42,46,7,10,19,18,18,20,10,7,11,13,
            36,25,33,24,28,30,38,31,48,40,61,32,30,44,52,39,45,47,40,44,
            43,42,44,34,52,45,40,58,55,41,52,40,62,49,36,40,48,58,41,42,
            37,41,59,42,50,52,35,52,44,38,53,65,48,47,57,53,43,52,32,49,
            19,18,17,17,15,18,12,18,12,8,53,57,42,47,42,41,49,51,45,44,
            49,47,53,33,36,37,44,40,70,57)
)

plot(schools_data, pch = 19, col = "red", main = "SIS School Closure Data")
```

### Model Definition

```{r}
sis <- odin({
  update(S) <- S - n_SI + n_IS
  update(I) <- I + n_SI - n_IS
  update(incidence) <- incidence + n_SI

  initial(S) <- N - I0
  initial(I) <- I0
  initial(incidence, zero_every = 1) <- 0

  schools <- interpolate(schools_time, schools_open, "constant")
  schools_time[] <- parameter()
  schools_open[] <- parameter()
  dim(schools_time, schools_open) <- parameter(rank = 1)

  beta <- ((1 - schools) * (1 - schools_modifier) + schools) * beta0

  p_SI <- 1 - exp(-beta * I / N * dt)
  p_IS <- 1 - exp(-gamma * dt)
  n_SI <- Binomial(S, p_SI)
  n_IS <- Binomial(I, p_IS)

  N <- parameter(1000)
  I0 <- parameter(10)
  beta0 <- parameter(0.2)
  gamma <- parameter(0.1)
  schools_modifier <- parameter(0.6)

  cases <- data()
  cases ~ Poisson(incidence)
})
```

### School Schedule

```{r}
schools_time <- c(0, 50, 60, 120, 130, 170, 180)
schools_open <- c(1,  0,  1,   0,   1,   0,   1)
```

### MCMC Setup

```{r}
sis_packer <- monty_packer(c("beta0", "gamma", "schools_modifier"),
                           fixed = list(schools_time = schools_time,
                                        schools_open = schools_open))

sis_filter <- dust_filter_create(sis, time_start = 0, dt = 1,
                                  data = schools_data, n_particles = 200)

sis_prior <- monty_dsl({
  beta0 ~ Exponential(mean = 0.3)
  gamma ~ Exponential(mean = 0.1)
  schools_modifier ~ Uniform(0, 1)
})

sis_likelihood <- dust_likelihood_monty(sis_filter, sis_packer)
sis_posterior <- sis_likelihood + sis_prior
```

### Run MCMC (4 chains)

```{r}
sis_vcv <- diag(c(2e-4, 2e-4, 4e-4))
sis_sampler <- monty_sampler_random_walk(sis_vcv)

sis_samples <- monty_sample(sis_posterior, sis_sampler, 500,
                            initial = c(0.3, 0.1, 0.5),
                            n_chains = 4)
```

### Posterior Summary

```{r}
sis_df <- posterior::as_draws_df(sis_samples)
posterior::summarise_draws(sis_df)
```

### Trace Plots

```{r}
bayesplot::mcmc_trace(sis_df)
```

### Log-Density Trace

```{r}
matplot(sis_samples$density, type = "l", lty = 1,
        xlab = "Iteration", ylab = "Log-density",
        main = "SIS Log-posterior trace")
```

## Part 3: Counterfactual Projections

### Forward Projection

Simulate from the posterior forward to day 200:

```{r}
sis_samples_thin <- monty_samples_thin(sis_samples,
                                        burnin = 100,
                                        thinning_factor = 4)
n_draws <- prod(dim(sis_samples_thin$pars)[-1])
pars_arr <- array(sis_samples_thin$pars, c(3, n_draws))

# Simulate each draw
projection_inc <- matrix(NA, nrow = 201, ncol = n_draws)
for (i in seq_len(n_draws)) {
  p <- sis_packer$unpack(pars_arr[, i])
  sys <- dust_system_create(sis, p, dt = 1)
  dust_system_set_state_initial(sys)
  t_all <- seq(0, 200)
  y <- dust_system_simulate(sys, t_all)
  y_unpacked <- dust_unpack_state(sys, y)
  projection_inc[, i] <- y_unpacked$incidence
}

matplot(0:200, projection_inc, type = "l", col = "#00000011", lty = 1,
        xlab = "Time", ylab = "Incidence",
        main = "Forward Projection")
points(schools_data, pch = 19, col = "red", cex = 0.5)
abline(v = 150, lty = 2)
```

### Counterfactual: Only First School Closure

```{r}
cf_inc <- matrix(NA, nrow = 201, ncol = n_draws)
for (i in seq_len(n_draws)) {
  p <- sis_packer$unpack(pars_arr[, i])
  p$schools_time <- c(0, 50, 60)
  p$schools_open <- c(1, 0, 1)
  sys <- dust_system_create(sis, p, dt = 1)
  dust_system_set_state_initial(sys)
  y <- dust_system_simulate(sys, seq(0, 200))
  y_unpacked <- dust_unpack_state(sys, y)
  cf_inc[, i] <- y_unpacked$incidence
}

matplot(0:200, projection_inc, type = "l", col = "#00000011", lty = 1,
        xlab = "Time", ylab = "Incidence",
        main = "Counterfactual: Only First Closure")
matlines(0:200, cf_inc, col = "#0000FF11", lty = 1)
points(schools_data, pch = 19, col = "red", cex = 0.5)
legend("topright", c("Fitted (3 closures)", "Only 1st closure"),
       col = c("black", "blue"), lty = 1)
```

### Counterfactual: No School Closures

```{r}
no_closure_inc <- matrix(NA, nrow = 201, ncol = n_draws)
for (i in seq_len(n_draws)) {
  p <- sis_packer$unpack(pars_arr[, i])
  p$schools_time <- 0
  p$schools_open <- 1
  sys <- dust_system_create(sis, p, dt = 1)
  dust_system_set_state_initial(sys)
  y <- dust_system_simulate(sys, seq(0, 200))
  y_unpacked <- dust_unpack_state(sys, y)
  no_closure_inc[, i] <- y_unpacked$incidence
}

matplot(0:200, projection_inc, type = "l", col = "#00000011", lty = 1,
        xlab = "Time", ylab = "Incidence",
        main = "Counterfactual: No Closures")
matlines(0:200, no_closure_inc, col = "#FFA50011", lty = 1)
points(schools_data, pch = 19, col = "red", cex = 0.5)
legend("topright", c("Fitted (3 closures)", "No closures"),
       col = c("black", "orange"), lty = 1)
```

## Summary

| Step | R API |
|------|-------|
| Define model | `odin({...})` |
| Deterministic likelihood | `dust_unfilter_create()` |
| Stochastic likelihood | `dust_filter_create()` |
| Bridge to MCMC | `dust_likelihood_monty()` |
| Define prior | `monty_dsl({...})` |
| Combine | `posterior <- likelihood + prior` |
| Sample | `monty_sample()` |
| Thin | `monty_samples_thin()` |
| Counterfactuals | `dust_system_create()` + `dust_system_simulate()` |
