---
title: "Time-Varying Parameters"
subtitle: "odin2/dust2 Vignette 08 (R)"
format:
  html:
    toc: true
    code-fold: false
  gfm: default
---

## Introduction

Matching the Julia vignette: time-varying transmission rate using odin2's `interpolate()`.

## Constant Interpolation (Step Function)

```{r}
library(odin2)
library(dust2)

sir_step <- odin({
  deriv(S) <- -beta * S * I / N
  deriv(I) <- beta * S * I / N - gamma * I
  deriv(R) <- gamma * I
  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0

  beta <- interpolate(beta_time, beta_value, "constant")
  beta_time[] <- parameter()
  beta_value[] <- parameter()
  dim(beta_time, beta_value) <- parameter(rank = 1)
  gamma <- parameter(0.1)
  I0 <- parameter(10)
  N <- parameter(1000)
})

pars <- list(
  beta_time = c(0, 30, 60),
  beta_value = c(0.5, 0.15, 0.3),
  gamma = 0.1, I0 = 10, N = 1000
)

sys <- dust_system_create(sir_step, pars, ode_control = dust_ode_control())
dust_system_set_state_initial(sys)
times <- seq(0, 120, by = 0.5)
result <- dust_system_simulate(sys, times)

plot(times, result[1, ], type = "l", col = "blue", lwd = 2,
     ylim = c(0, 1000), xlab = "Time", ylab = "Population",
     main = "SIR with Step-Function β")
lines(times, result[2, ], col = "red", lwd = 2)
lines(times, result[3, ], col = "green", lwd = 2)
abline(v = c(30, 60), lty = 2, col = "gray")
legend("right", legend = c("S", "I", "R"), col = c("blue", "red", "green"), lwd = 2)
```

## Linear Interpolation

```{r}
sir_linear <- odin({
  deriv(S) <- -beta * S * I / N
  deriv(I) <- beta * S * I / N - gamma * I
  deriv(R) <- gamma * I
  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0

  beta <- interpolate(beta_time, beta_value, "linear")
  beta_time[] <- parameter()
  beta_value[] <- parameter()
  dim(beta_time, beta_value) <- parameter(rank = 1)
  gamma <- parameter(0.1)
  I0 <- parameter(10)
  N <- parameter(1000)
})

pars_linear <- list(
  beta_time = c(0, 40, 80, 120),
  beta_value = c(0.5, 0.3, 0.1, 0.2),
  gamma = 0.1, I0 = 10, N = 1000
)

sys_l <- dust_system_create(sir_linear, pars_linear, ode_control = dust_ode_control())
dust_system_set_state_initial(sys_l)
result_l <- dust_system_simulate(sys_l, times)

plot(times, result_l[2, ], type = "l", col = "red", lwd = 2,
     xlab = "Time", ylab = "Infected",
     main = "SIR with Linearly Varying β")
```

## Comparing Step and Linear

```{r}
plot(times, result[2, ], type = "l", col = "blue", lwd = 2,
     xlab = "Time", ylab = "Infected",
     main = "Effect of Interpolation Mode")
lines(times, result_l[2, ], col = "red", lwd = 2, lty = 2)
legend("topright", legend = c("Step", "Linear"), col = c("blue", "red"), lwd = 2, lty = 1:2)
```

## Summary

Both Julia and R support interpolated time-varying parameters. R's odin2 uses
string mode names (`"constant"`, `"linear"`, `"spline"`), while Julia uses
symbols (`:constant`, `:linear`, `:spline`).
