---
title: "Deterministic Mpox Inference Benchmark (R)"
subtitle: "odin2/dust2/monty Companion to Odin.jl Vignette 38"
format:
  html:
    toc: true
    code-fold: false
  gfm: default
---

## Introduction

This companion vignette mirrors the Julia deterministic mpox benchmark with the
same compact intervention model and the same synthetic data recipe. It focuses
on two quantities that are directly comparable across the two ecosystems:

1. deterministic trajectory simulation,
2. deterministic likelihood evaluation.

For gradients, this R version uses a simple finite-difference benchmark as a
baseline.

```{r}
suppressPackageStartupMessages({
  library(odin2)
  library(dust2)
  library(monty)
})
```

## Model definition

```{r}
mpox_det <- odin({
  total_I <- I_mild + I_hosp + I_ICU
  beta <- if (time < intervention_day) beta_1 else beta_2
  contact <- if (time < intervention_day) contact_1 else contact_2

  deriv(S) <- -beta * contact * S * total_I / N
  deriv(E) <- beta * contact * S * total_I / N - gamma * E
  deriv(I_mild) <- p_mild * gamma * E - sigma * I_mild - mu * I_mild
  deriv(I_hosp) <- p_hosp * gamma * E - sigma * I_hosp - mu * I_hosp
  deriv(I_ICU) <- p_ICU * gamma * E - sigma * I_ICU - mu * I_ICU
  deriv(R) <- sigma * total_I
  deriv(D) <- mu * total_I

  N <- S + E + I_mild + I_hosp + I_ICU + R + D

  cases <- data()
  cases ~ Poisson(max(total_I, 1e-6))

  initial(S) <- S0
  initial(E) <- E0
  initial(I_mild) <- I_mild0
  initial(I_hosp) <- I_hosp0
  initial(I_ICU) <- I_ICU0
  initial(R) <- R0
  initial(D) <- D0

  S0 <- parameter(9969)
  E0 <- parameter(20)
  I_mild0 <- parameter(8)
  I_hosp0 <- parameter(2)
  I_ICU0 <- parameter(1)
  R0 <- parameter(0)
  D0 <- parameter(0)

  gamma <- parameter(1 / 10)
  sigma <- parameter(1 / 7)
  mu <- parameter(0.01)
  p_mild <- parameter(0.85)
  p_hosp <- parameter(0.10)
  p_ICU <- parameter(0.05)

  beta_1 <- parameter(0.9)
  beta_2 <- parameter(0.45)
  contact_1 <- parameter(1.0)
  contact_2 <- parameter(0.65)
  intervention_day <- parameter(40)
})
```

## Simulation setup

```{r}
true_pars <- list(
  S0 = 9969, E0 = 20, I_mild0 = 8, I_hosp0 = 2, I_ICU0 = 1, R0 = 0, D0 = 0,
  gamma = 1 / 10, sigma = 1 / 7, mu = 0.01, p_mild = 0.85, p_hosp = 0.10, p_ICU = 0.05,
  beta_1 = 0.9, beta_2 = 0.45, contact_1 = 1.0, contact_2 = 0.65, intervention_day = 40
)

times <- 0:90
sys <- dust_system_create(
  mpox_det,
  true_pars,
  n_particles = 1L,
  preserve_particle_dimension = TRUE
)
dust_system_set_state_initial(sys)
sim <- dust_system_simulate(sys, times)

obs_times <- seq(10, 80, by = 5)
obs_idx <- match(obs_times, times)

set.seed(99)
obs_cases <- sapply(obs_idx, function(i) {
  rpois(1, lambda = max(sum(sim[3:5, 1, i]), 1e-6))
})

obs <- data.frame(time = obs_times, cases = obs_cases)
obs
```

## Deterministic likelihood

```{r}
candidate_pars <- modifyList(true_pars, list(
  beta_1 = 0.82,
  beta_2 = 0.52,
  contact_1 = 1.05,
  contact_2 = 0.60
))

lik <- dust_unfilter_create(mpox_det, time_start = 0, data = obs)
ll <- dust_likelihood_run(lik, candidate_pars)
ll
```

## Finite-difference gradient baseline

```{r}
fd_gradient <- function(pars, eps = 1e-5) {
  target <- c("beta_1", "beta_2", "contact_1", "contact_2")
  g <- numeric(length(target))
  names(g) <- target

  for (i in seq_along(target)) {
    nm <- target[i]
    p_plus <- pars
    p_minus <- pars
    p_plus[[nm]] <- p_plus[[nm]] + eps
    p_minus[[nm]] <- p_minus[[nm]] - eps
    g[i] <- (
      dust_likelihood_run(lik, p_plus) -
      dust_likelihood_run(lik, p_minus)
    ) / (2 * eps)
  }

  g
}

fd_gradient(candidate_pars)
```

## Benchmark summary

```{r}
simulation_ms <- system.time(replicate(20, {
  s <- dust_system_create(
    mpox_det,
    true_pars,
    n_particles = 1L,
    preserve_particle_dimension = TRUE
  )
  dust_system_set_state_initial(s)
  dust_system_simulate(s, times)
}))[["elapsed"]] / 20 * 1000

likelihood_ms <- system.time(replicate(20, {
  dust_likelihood_run(lik, candidate_pars)
}))[["elapsed"]] / 20 * 1000

fd_ms <- system.time(replicate(5, {
  fd_gradient(candidate_pars)
}))[["elapsed"]] / 5 * 1000

data.frame(
  simulation_ms = simulation_ms,
  likelihood_ms = likelihood_ms,
  finite_difference_gradient_ms = fd_ms,
  fd_over_likelihood = fd_ms / likelihood_ms
)
```

## DZA real-data slice

We also vendor a processed slice of the DZA example data plus the associated
intervention multiplier into this vignette directory so the benchmark can run
without depending on the external `mpoxseir` checkout.

```{r}
dza_daily <- read.csv(file.path("..", "data", "dza_daily.csv"))
dza_intervention <- read.csv(file.path("..", "data", "dza_intervention.csv"))

obs_dza <- data.frame(
  time = dza_daily$day,
  cases = dza_daily$cases,
  deaths = dza_daily$deaths
)

summary(obs_dza)
```

## DZA interpolation model

```{r}
mpox_det_dza <- odin({
  contact_multiplier <- interpolate(contact_time, contact_value, "constant")
  total_I <- I_mild + I_hosp + I_ICU
  force <- beta * contact_multiplier * S * total_I / N
  cases_rate <- case_scale * gamma * E
  deaths_rate <- death_scale * mu * total_I

  deriv(S) <- -force
  deriv(E) <- force - gamma * E
  deriv(I_mild) <- p_mild * gamma * E - sigma * I_mild - mu * I_mild
  deriv(I_hosp) <- p_hosp * gamma * E - sigma * I_hosp - mu * I_hosp
  deriv(I_ICU) <- p_ICU * gamma * E - sigma * I_ICU - mu * I_ICU
  deriv(R) <- sigma * total_I
  deriv(D) <- deaths_rate

  N <- S + E + I_mild + I_hosp + I_ICU + R + D

  cases <- data()
  deaths <- data()
  cases ~ Poisson(max(cases_rate, 1e-6))
  deaths ~ Poisson(max(deaths_rate, 1e-6))

  initial(S) <- S0
  initial(E) <- E0
  initial(I_mild) <- I_mild0
  initial(I_hosp) <- I_hosp0
  initial(I_ICU) <- I_ICU0
  initial(R) <- R0
  initial(D) <- D0

  contact_time[] <- parameter()
  contact_value[] <- parameter()
  dim(contact_time, contact_value) <- parameter(rank = 1)

  S0 <- parameter(500000)
  E0 <- parameter(40)
  I_mild0 <- parameter(10)
  I_hosp0 <- parameter(3)
  I_ICU0 <- parameter(1)
  R0 <- parameter(0)
  D0 <- parameter(0)
  gamma <- parameter(1 / 7)
  sigma <- parameter(1 / 10)
  mu <- parameter(0.004)
  p_mild <- parameter(0.85)
  p_hosp <- parameter(0.10)
  p_ICU <- parameter(0.05)
  beta <- parameter(0.38)
  case_scale <- parameter(8)
  death_scale <- parameter(2.5)
})
```

## DZA likelihood and finite-difference benchmark

```{r}
dza_pars <- list(
  contact_time = dza_intervention$day,
  contact_value = dza_intervention$contact_multiplier,
  S0 = 500000, E0 = 40, I_mild0 = 10, I_hosp0 = 3, I_ICU0 = 1, R0 = 0, D0 = 0,
  gamma = 1 / 7, sigma = 1 / 10, mu = 0.004, p_mild = 0.85, p_hosp = 0.10, p_ICU = 0.05,
  beta = 0.38, case_scale = 8, death_scale = 2.5
)

dza_lik <- dust_unfilter_create(mpox_det_dza, time_start = -1, data = obs_dza)
dza_ll <- dust_likelihood_run(dza_lik, dza_pars)
dza_ll
```

```{r}
fd_gradient_dza <- function(pars, eps = 1e-5) {
  target <- c("beta", "mu", "case_scale", "death_scale")
  g <- numeric(length(target))
  names(g) <- target

  for (i in seq_along(target)) {
    nm <- target[i]
    p_plus <- pars
    p_minus <- pars
    p_plus[[nm]] <- p_plus[[nm]] + eps
    p_minus[[nm]] <- p_minus[[nm]] - eps
    g[i] <- (
      dust_likelihood_run(dza_lik, p_plus) -
      dust_likelihood_run(dza_lik, p_minus)
    ) / (2 * eps)
  }

  g
}

fd_gradient_dza(dza_pars)
```

```{r}
dza_simulation_ms <- system.time(replicate(20, {
  s <- dust_system_create(
    mpox_det_dza,
    dza_pars,
    n_particles = 1L,
    preserve_particle_dimension = TRUE
  )
  dust_system_set_state_initial(s)
  dust_system_simulate(s, 0:max(dza_daily$day))
}))[["elapsed"]] / 20 * 1000

dza_likelihood_ms <- system.time(replicate(20, {
  dust_likelihood_run(dza_lik, dza_pars)
}))[["elapsed"]] / 20 * 1000

dza_fd_ms <- system.time(replicate(5, {
  fd_gradient_dza(dza_pars)
}))[["elapsed"]] / 5 * 1000

data.frame(
  simulation_ms = dza_simulation_ms,
  likelihood_ms = dza_likelihood_ms,
  finite_difference_gradient_ms = dza_fd_ms,
  fd_over_likelihood = dza_fd_ms / dza_likelihood_ms
)
```

## DZA point fit

For a small deterministic fitting step, we use the same log-space backtracking
ascent idea as the Julia vignette. Here the gradient comes from the finite-
difference baseline, which makes the runtime gap between the two ecosystems
explicit while still keeping the overall optimisation logic comparable.

```{r}
fit_logspace_fd <- function(lik, fixed_pars, initial_free,
                            target = c("beta", "case_scale", "death_scale"),
                            step = 0.2, max_iter = 40, min_step = 1e-6,
                            eps = 1e-5) {
  theta <- log(unlist(initial_free[target], use.names = FALSE))
  names(theta) <- target

  current_free <- as.list(stats::setNames(exp(theta), target))
  current_pars <- modifyList(fixed_pars, current_free)
  current_ll <- dust_likelihood_run(lik, current_pars)
  current_grad <- fd_gradient_dza(current_pars, eps = eps)[target]
  history <- current_ll

  for (iter in seq_len(max_iter)) {
    g_theta <- as.numeric(current_grad) * exp(theta)
    grad_norm <- sqrt(sum(g_theta^2))
    if (grad_norm == 0) {
      break
    }

    accepted <- FALSE
    trial_step <- step
    while (trial_step >= min_step) {
      theta_trial <- theta + trial_step * g_theta / max(grad_norm, 1)
      free_trial <- as.list(stats::setNames(exp(theta_trial), target))
      pars_trial <- modifyList(fixed_pars, free_trial)
      ll_trial <- dust_likelihood_run(lik, pars_trial)

      if (is.finite(ll_trial) && ll_trial > current_ll + 1e-6) {
        theta <- theta_trial
        current_free <- free_trial
        current_pars <- pars_trial
        current_ll <- ll_trial
        current_grad <- fd_gradient_dza(current_pars, eps = eps)[target]
        history <- c(history, current_ll)
        accepted <- TRUE
        break
      }

      trial_step <- trial_step / 2
    }

    if (!accepted) {
      break
    }
  }

  list(
    free = current_free,
    pars = current_pars,
    log_likelihood = current_ll,
    history = history
  )
}

dza_fit_fixed <- modifyList(
  dza_pars,
  list(beta = NULL, case_scale = NULL, death_scale = NULL)
)
dza_fit_initial <- list(beta = 0.24, case_scale = 5.5, death_scale = 1.6)
dza_fit_start_ll <- dust_likelihood_run(dza_lik, modifyList(dza_fit_fixed, dza_fit_initial))
dza_fit <- fit_logspace_fd(dza_lik, dza_fit_fixed, dza_fit_initial)
dza_fit_ms <- system.time({
  fit_logspace_fd(dza_lik, dza_fit_fixed, dza_fit_initial)
})[["elapsed"]] * 1000

list(
  start_loglik = dza_fit_start_ll,
  final_loglik = dza_fit$log_likelihood,
  fitted = dza_fit$free,
  n_iterations = length(dza_fit$history) - 1,
  point_fit_ms = dza_fit_ms
)
```

## Summary

The R and Julia benchmark vignettes now share both a compact synthetic mpox
benchmark and a processed DZA real-data slice. In Julia, this is enough to
compare forward and adjoint gradients directly. In the R companion, the
matching baseline is finite differences, which makes the gradient cost gap
visible on both the synthetic and real-data deterministic examples, including a
small DZA point-fit workflow.

## Age-structured extension

This section adds a 3-age-group SEIRD model that matches the Julia vignette's
age-structured slice. The model uses split latent stages, age-specific
susceptibility and CFR, and per-age Poisson case/death observations.

```{r}
mpox_age <- odin({
    n_age <- parameter(3)

    dim(susc) <- n_age
    susc[] <- parameter()
    dim(cfr) <- n_age
    cfr[] <- parameter()

    dim(I_total) <- n_age
    I_total[] <- Ir[i] + Id[i]
    total_I <- sum(I_total)
    dim(lambda) <- n_age
    lambda[] <- beta * susc[i] * total_I / N

    dim(S) <- n_age
    dim(Ea) <- n_age
    dim(Eb) <- n_age
    dim(Ir) <- n_age
    dim(Id) <- n_age
    dim(R) <- n_age
    dim(D) <- n_age

    deriv(S[])  <- -lambda[i] * S[i]
    deriv(Ea[]) <-  lambda[i] * S[i] - 2 * gamma_E * Ea[i]
    deriv(Eb[]) <-  2 * gamma_E * Ea[i] - 2 * gamma_E * Eb[i]
    deriv(Ir[]) <- (1 - cfr[i]) * 2 * gamma_E * Eb[i] - sigma * Ir[i]
    deriv(Id[]) <-  cfr[i]      * 2 * gamma_E * Eb[i] - mu * Id[i]
    deriv(R[])  <-  sigma * Ir[i]
    deriv(D[])  <-  mu * Id[i]

    N <- sum(S) + sum(Ea) + sum(Eb) + sum(Ir) + sum(Id) + sum(R) + sum(D)

    dim(cases_rate) <- n_age
    dim(deaths_rate) <- n_age
    cases_rate[]  <- case_scale  * 2 * gamma_E * Eb[i]
    deaths_rate[] <- death_scale * mu * Id[i]

    dim(cases) <- n_age
    dim(deaths) <- n_age
    cases[] <- data()
    deaths[] <- data()
    cases[] ~ Poisson(max(cases_rate[i], 1e-6))
    deaths[] ~ Poisson(max(deaths_rate[i], 1e-6))

    dim(S0) <- n_age
    dim(Ea0) <- n_age
    S0[] <- parameter()
    Ea0[] <- parameter()
    initial(S[])  <- S0[i]
    initial(Ea[]) <- Ea0[i]
    initial(Eb[]) <- 0
    initial(Ir[]) <- 0
    initial(Id[]) <- 0
    initial(R[])  <- 0
    initial(D[])  <- 0

    beta        <- parameter(0.065)
    gamma_E     <- parameter(1 / 10)
    sigma       <- parameter(1 / 14)
    mu          <- parameter(1 / 7)
    case_scale  <- parameter(6.0)
    death_scale <- parameter(2.0)
})
```

### Simulation and synthetic data

```{r}
age_pars <- list(
    n_age = 3,
    susc = c(1.0, 1.2, 0.8),
    cfr  = c(0.005, 0.02, 0.08),
    S0   = c(4000, 3500, 2500),
    Ea0  = c(8, 6, 4),
    beta = 0.065, gamma_E = 1/10, sigma = 1/14, mu = 1/7,
    case_scale = 6.0, death_scale = 2.0
)

age_sys <- dust_system_create(mpox_age, age_pars, deterministic = TRUE)
age_sim_times <- seq(0, 120, by = 1)
dust_system_set_state_initial(age_sys)
age_sim <- dust_system_simulate(age_sys, age_sim_times)
dim(age_sim)
```

```{r}
age_state_names <- rownames(age_sim)
idx_Eb <- grep("^Eb\\[", age_state_names)
idx_Id <- grep("^Id\\[", age_state_names)
cat("Eb indices:", idx_Eb, "\nId indices:", idx_Id, "\n")
cat("age_sim dimensions:", paste(dim(age_sim), collapse=" x "), "\n")

set.seed(77)
age_obs_times <- seq(7, 112, by = 7)

# Handle both 2D (state × time) and 3D (state × particle × time) output
get_state <- function(sim, rows, ti) {
    if (length(dim(sim)) == 3) {
        sim[rows, 1, ti]
    } else {
        sim[rows, ti]
    }
}

age_obs_list <- lapply(age_obs_times, function(t) {
    ti <- which(age_sim_times == t)
    c_rate <- age_pars$case_scale * 2 * age_pars$gamma_E * get_state(age_sim, idx_Eb, ti)
    d_rate <- age_pars$death_scale * age_pars$mu * get_state(age_sim, idx_Id, ti)
    list(
        cases  = pmax(1, rpois(3, pmax(c_rate, 1e-6))),
        deaths = pmax(1, rpois(3, pmax(d_rate, 1e-6)))
    )
})

age_data <- data.frame(
    time   = age_obs_times,
    cases  = I(lapply(age_obs_list, function(x) x$cases)),
    deaths = I(lapply(age_obs_list, function(x) x$deaths))
)
head(age_data)
```

### Deterministic likelihood

```{r}
age_filter <- dust_unfilter_create(mpox_age, age_data, time_start = 0)
age_ll <- dust_likelihood_run(age_filter, age_pars)
cat("Age-structured log-likelihood:", age_ll, "\n")

age_ll_ms <- median(replicate(10, {
    system.time(dust_likelihood_run(age_filter, age_pars))[["elapsed"]]
})) * 1000
cat("Loglik median time:", round(age_ll_ms, 2), "ms\n")
```

### Finite-difference gradients

```{r}
age_fd_gradient <- function(lik, pars, param_names, eps = 1e-5) {
    base_ll <- dust_likelihood_run(lik, pars)
    grad <- numeric(length(param_names))
    for (j in seq_along(param_names)) {
        pars_plus <- pars
        pars_plus[[param_names[j]]] <- pars_plus[[param_names[j]]] + eps
        grad[j] <- (dust_likelihood_run(lik, pars_plus) - base_ll) / eps
    }
    list(gradient = grad, log_likelihood = base_ll)
}

age_fd <- age_fd_gradient(age_filter, age_pars,
                          c("beta", "case_scale", "death_scale"))
cat("FD gradient:", round(age_fd$gradient, 4), "\n")

age_fd_ms <- median(replicate(5, {
    system.time(age_fd_gradient(age_filter, age_pars,
        c("beta", "case_scale", "death_scale")))[["elapsed"]]
})) * 1000
cat("FD gradient median time:", round(age_fd_ms, 2), "ms\n")
```

### Point fit

```{r}
age_fit_fixed <- modifyList(age_pars, list(
    beta = NULL, case_scale = NULL, death_scale = NULL
))
age_fit_initial <- list(beta = 0.04, case_scale = 4.0, death_scale = 1.0)
age_fit_start_ll <- dust_likelihood_run(age_filter,
    modifyList(age_fit_fixed, age_fit_initial))

# Simple FD gradient ascent in log-space
age_fit_logspace <- function(lik, fixed, initial,
                             step = 0.15, max_iter = 30, min_step = 1e-6) {
    theta <- log(unlist(initial))
    pnames <- names(initial)
    current_pars <- modifyList(fixed, as.list(exp(theta)))
    names(current_pars[[pnames[1]]]) <- NULL
    current_ll <- dust_likelihood_run(lik, current_pars)
    history <- current_ll
    eps <- 1e-5

    for (iter in seq_len(max_iter)) {
        grad_raw <- numeric(length(theta))
        for (j in seq_along(pnames)) {
            tp <- theta
            tp[j] <- tp[j] + eps
            pp <- modifyList(fixed, as.list(setNames(exp(tp), pnames)))
            grad_raw[j] <- (dust_likelihood_run(lik, pp) - current_ll) / eps
        }
        gnorm <- sqrt(sum(grad_raw^2))
        if (gnorm == 0) break

        accepted <- FALSE
        trial <- step
        while (trial >= min_step) {
            tt <- theta + trial * grad_raw / max(gnorm, 1)
            pp <- modifyList(fixed, as.list(setNames(exp(tt), pnames)))
            ll <- dust_likelihood_run(lik, pp)
            if (is.finite(ll) && ll > current_ll + 1e-6) {
                theta <- tt
                current_pars <- pp
                current_ll <- ll
                history <- c(history, ll)
                accepted <- TRUE
                break
            }
            trial <- trial * 0.5
        }
        if (!accepted) break
    }

    list(
        free = as.list(setNames(exp(theta), pnames)),
        log_likelihood = current_ll,
        history = history
    )
}

age_fit <- age_fit_logspace(age_filter, age_fit_fixed, age_fit_initial)

list(
    start_loglik = age_fit_start_ll,
    final_loglik = age_fit$log_likelihood,
    fitted       = age_fit$free,
    improvement  = age_fit$log_likelihood - age_fit_start_ll,
    n_iterations = length(age_fit$history) - 1
)
```

## Full summary

This vignette covers three progressively richer deterministic mpox models:

1. **Compact synthetic** — scalar SEIRD with piecewise intervention,
2. **DZA real-data slice** — scalar SEIRD with interpolated contact multiplier
   and real weekly observations,
3. **Age-structured slice** — 3-age-group SEIRD with split latent stages,
   age-specific susceptibility/CFR, and per-age Poisson case/death streams.

The R companion matches the Julia vignette layout, using finite-difference
gradients where Julia uses forward/adjoint analytical gradients.
