---
title: "Deterministic Mpox Inference Benchmark (R)"
subtitle: "odin2/dust2/monty Companion to Odin.jl Vignette 38"
format:
  html:
    toc: true
    code-fold: false
  gfm: default
---

## Introduction

This companion vignette mirrors the Julia deterministic mpox benchmark with the
same compact intervention model and the same synthetic data recipe. It focuses
on two quantities that are directly comparable across the two ecosystems:

1. deterministic trajectory simulation,
2. deterministic likelihood evaluation.

For gradients, this R version uses a simple finite-difference benchmark as a
baseline.

```{r}
suppressPackageStartupMessages({
  library(odin2)
  library(dust2)
  library(monty)
})
```

## Model definition

```{r}
mpox_det <- odin({
  total_I <- I_mild + I_hosp + I_ICU
  beta <- if (time < intervention_day) beta_1 else beta_2
  contact <- if (time < intervention_day) contact_1 else contact_2

  deriv(S) <- -beta * contact * S * total_I / N
  deriv(E) <- beta * contact * S * total_I / N - gamma * E
  deriv(I_mild) <- p_mild * gamma * E - sigma * I_mild - mu * I_mild
  deriv(I_hosp) <- p_hosp * gamma * E - sigma * I_hosp - mu * I_hosp
  deriv(I_ICU) <- p_ICU * gamma * E - sigma * I_ICU - mu * I_ICU
  deriv(R) <- sigma * total_I
  deriv(D) <- mu * total_I

  N <- S + E + I_mild + I_hosp + I_ICU + R + D

  cases <- data()
  cases ~ Poisson(max(total_I, 1e-6))

  initial(S) <- S0
  initial(E) <- E0
  initial(I_mild) <- I_mild0
  initial(I_hosp) <- I_hosp0
  initial(I_ICU) <- I_ICU0
  initial(R) <- R0
  initial(D) <- D0

  S0 <- parameter(9969)
  E0 <- parameter(20)
  I_mild0 <- parameter(8)
  I_hosp0 <- parameter(2)
  I_ICU0 <- parameter(1)
  R0 <- parameter(0)
  D0 <- parameter(0)

  gamma <- parameter(1 / 10)
  sigma <- parameter(1 / 7)
  mu <- parameter(0.01)
  p_mild <- parameter(0.85)
  p_hosp <- parameter(0.10)
  p_ICU <- parameter(0.05)

  beta_1 <- parameter(0.9)
  beta_2 <- parameter(0.45)
  contact_1 <- parameter(1.0)
  contact_2 <- parameter(0.65)
  intervention_day <- parameter(40)
})
```

## Simulation setup

```{r}
true_pars <- list(
  S0 = 9969, E0 = 20, I_mild0 = 8, I_hosp0 = 2, I_ICU0 = 1, R0 = 0, D0 = 0,
  gamma = 1 / 10, sigma = 1 / 7, mu = 0.01, p_mild = 0.85, p_hosp = 0.10, p_ICU = 0.05,
  beta_1 = 0.9, beta_2 = 0.45, contact_1 = 1.0, contact_2 = 0.65, intervention_day = 40
)

times <- 0:90
sys <- dust_system_create(
  mpox_det,
  true_pars,
  n_particles = 1L,
  preserve_particle_dimension = TRUE
)
dust_system_set_state_initial(sys)
sim <- dust_system_simulate(sys, times)

obs_times <- seq(10, 80, by = 5)
obs_idx <- match(obs_times, times)

set.seed(99)
obs_cases <- sapply(obs_idx, function(i) {
  rpois(1, lambda = max(sum(sim[3:5, 1, i]), 1e-6))
})

obs <- data.frame(time = obs_times, cases = obs_cases)
obs
```

## Deterministic likelihood

```{r}
candidate_pars <- modifyList(true_pars, list(
  beta_1 = 0.82,
  beta_2 = 0.52,
  contact_1 = 1.05,
  contact_2 = 0.60
))

lik <- dust_unfilter_create(mpox_det, time_start = 0, data = obs)
ll <- dust_likelihood_run(lik, candidate_pars)
ll
```

## Finite-difference gradient baseline

```{r}
fd_gradient <- function(pars, eps = 1e-5) {
  target <- c("beta_1", "beta_2", "contact_1", "contact_2")
  g <- numeric(length(target))
  names(g) <- target

  for (i in seq_along(target)) {
    nm <- target[i]
    p_plus <- pars
    p_minus <- pars
    p_plus[[nm]] <- p_plus[[nm]] + eps
    p_minus[[nm]] <- p_minus[[nm]] - eps
    g[i] <- (
      dust_likelihood_run(lik, p_plus) -
      dust_likelihood_run(lik, p_minus)
    ) / (2 * eps)
  }

  g
}

fd_gradient(candidate_pars)
```

## Benchmark summary

```{r}
simulation_ms <- system.time(replicate(20, {
  s <- dust_system_create(
    mpox_det,
    true_pars,
    n_particles = 1L,
    preserve_particle_dimension = TRUE
  )
  dust_system_set_state_initial(s)
  dust_system_simulate(s, times)
}))[["elapsed"]] / 20 * 1000

likelihood_ms <- system.time(replicate(20, {
  dust_likelihood_run(lik, candidate_pars)
}))[["elapsed"]] / 20 * 1000

fd_ms <- system.time(replicate(5, {
  fd_gradient(candidate_pars)
}))[["elapsed"]] / 5 * 1000

data.frame(
  simulation_ms = simulation_ms,
  likelihood_ms = likelihood_ms,
  finite_difference_gradient_ms = fd_ms,
  fd_over_likelihood = fd_ms / likelihood_ms
)
```

## Summary

The R and Julia benchmark vignettes now use the same deterministic mpox model
and the same synthetic data construction. In Julia, this is enough to compare
forward and adjoint gradients directly. In the R companion, the matching
baseline is finite differences, which makes the gradient cost gap visible even
on a compact deterministic mpox example.
