---
title: "DynamicPPL Integration (R Equivalent)"
subtitle: "odin2/dust2/monty Vignette 13 (R)"
format:
  html:
    toc: true
    code-fold: false
---

## Introduction

This vignette shows the R equivalent of the Julia DynamicPPL integration workflow. In R, we use `monty_dsl()` for prior specification and `sample()` for inference, with `odin2`, `dust2`, and `monty`.

Note that R's `monty_dsl()` supports independent priors only. The hierarchical prior structures available via Julia's DynamicPPL integration (e.g. deriving β from R₀ and γ) are not directly expressible in `monty_dsl()` — this is one advantage of the Julia ecosystem.

## Model and Data

```{r}
library(odin2)
library(dust2)
library(monty)

sir <- odin({
  deriv(S) <- -beta * S * Inf2 / N
  deriv(Inf2) <- beta * S * Inf2 / N - gamma * Inf2
  deriv(R) <- gamma * Inf2
  initial(S) <- N - I0
  initial(Inf2) <- I0
  initial(R) <- 0

  cases <- data()
  cases ~ Poisson(Inf2 + 1e-6)

  beta <- parameter(0.5)
  gamma <- parameter(0.1)
  I0 <- parameter(10)
  N <- parameter(1000)
})
```

## Generate Synthetic Data

```{r}
true_pars <- list(beta = 0.5, gamma = 0.1, I0 = 10, N = 1000)
times <- seq(0, 50, by = 1)

sys <- System(sir, true_pars, seed = 1)
dust_system_set_state_initial(sys)
result <- simulate(sys, times)

true_infected <- result[2, -1]
observed <- round(pmax(true_infected, 0))

data <- data.frame(time = times[-1], cases = observed)

plot(data$time, data$cases, pch = 16, cex = 0.8,
     xlab = "Time", ylab = "Cases", main = "Synthetic Data")
```

## Set Up Inference

```{r}
uf <- Likelihood(sir, time_start = 0, data = data)
pk <- Packer(c("beta", "gamma"), fixed = list(I0 = 10, N = 1000))
likelihood <- as_model(uf, pk)

cat("Log-likelihood at true parameters:",
    monty_model_density(likelihood, c(0.5, 0.1)), "\n")
```

## Define Priors with `monty_dsl`

In R, `monty_dsl()` is the standard way to define priors. It supports independent distributions:

```{r}
prior <- monty_dsl({
  beta ~ Gamma(shape = 2, rate = 4)
  gamma ~ Gamma(shape = 2, rate = 20)
})
```

This is equivalent to the Julia `@prior` and `dppl_prior` approaches shown in the Julia vignette.

```{r}
posterior <- likelihood + prior
```

## Run MCMC

```{r}
vcv <- matrix(c(0.005, 0, 0, 0.001), 2, 2)
sampler <- adaptive_mh(vcv)

samples <- sample(posterior, sampler, 3000, initial = c(0.4, 0.08),
                        n_chains = 1, burnin = 500)
```

## Results

```{r}
beta_samples <- samples$pars[1, , 1]
gamma_samples <- samples$pars[2, , 1]

cat("β: mean =", round(mean(beta_samples), 3),
    ", 95% CI = [", round(quantile(beta_samples, 0.025), 3),
    ",", round(quantile(beta_samples, 0.975), 3), "]\n")
cat("γ: mean =", round(mean(gamma_samples), 3),
    ", 95% CI = [", round(quantile(gamma_samples, 0.025), 3),
    ",", round(quantile(gamma_samples, 0.975), 3), "]\n")
cat("True: β = 0.5, γ = 0.1\n")
```

```{r}
par(mfrow = c(1, 2))
hist(beta_samples, breaks = 30, main = "Posterior: β", xlab = "β",
     probability = TRUE, col = "steelblue")
abline(v = 0.5, col = "red", lwd = 2)

hist(gamma_samples, breaks = 30, main = "Posterior: γ", xlab = "γ",
     probability = TRUE, col = "steelblue")
abline(v = 0.1, col = "red", lwd = 2)
```

## Trace Plots

```{r}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
plot(beta_samples, type = "l", xlab = "Iteration", ylab = "β", main = "Trace: β")
abline(h = 0.5, col = "red")
plot(gamma_samples, type = "l", xlab = "Iteration", ylab = "γ", main = "Trace: γ")
abline(h = 0.1, col = "red")
```

## Comparison with Julia

The R workflow produces equivalent posteriors to the Julia `@prior` and `dppl_prior` approaches. The key differences are:

| Feature | R (`monty_dsl`) | Julia (`@prior`) | Julia (`dppl_prior`) |
|---------|-----------------|------------------------|----------------------|
| Independent priors | ✓ | ✓ | ✓ |
| Hierarchical priors | ✗ | ✗ | ✓ |
| Auto-differentiation | Via dust2 C++ | Via ForwardDiff.jl | Via ForwardDiff.jl |
| MCMCChains diagnostics | Manual | `to_chains()` | `to_chains()` |
| LogDensityProblems | N/A | `as_logdensity()` | `as_logdensity()` |

For models requiring hierarchical priors (e.g. R₀-based parameterisation, random effects), the Julia DynamicPPL integration provides capabilities not available in the R ecosystem.
