---
title: "Stochastic Differential Equations (SDE) (R)"
format:
  html: default
  gfm: default
---

## Introduction

Stochastic differential equations (SDEs) extend ordinary differential equations (ODEs) with continuous noise terms, modelling systems subject to random fluctuations. In the Itô formulation:

$$
dX = f(X, t)\,dt + \sigma(X, t)\,dW
$$

where $f$ is the drift (deterministic dynamics), $\sigma$ is the diffusion coefficient (noise intensity), and $dW$ is a Wiener process increment.

While Julia's Odin supports SDEs natively via `diffusion()`, in R we implement Euler-Maruyama and Milstein schemes manually alongside `deSolve` for the deterministic ODE baseline.

## SIR ODE with deSolve

We start with the deterministic SIR model using `deSolve`:

```{r}
library(deSolve)
set.seed(42)

sir_ode <- function(t, state, pars) {
  with(as.list(c(state, pars)), {
    dS <- -beta * S * I / N
    dI <- beta * S * I / N - gamma * I
    dR <- gamma * I
    list(c(dS, dI, dR))
  })
}

pars <- c(beta = 0.5, gamma = 0.1, N = 1000)
state0 <- c(S = 990, I = 10, R = 0)
times <- seq(0, 80, by = 0.5)

out_ode <- ode(y = state0, times = times, func = sir_ode, parms = pars)
cat("ODE I peak:", max(out_ode[, "I"]), "\n")
cat("ODE peak time:", out_ode[which.max(out_ode[, "I"]), "time"], "\n")
```

## SIR SDE via Euler-Maruyama

The Euler-Maruyama scheme approximates the SDE at each time step:

$$
X_{n+1} = X_n + f(X_n)\,\Delta t + \sigma(X_n)\,\sqrt{\Delta t}\,Z_n
$$

where $Z_n \sim N(0, 1)$.

```{r}
sir_sde_euler_maruyama <- function(pars, state0, times, dt = 0.01, seed = 42) {
  set.seed(seed)
  beta <- pars["beta"]; gamma <- pars["gamma"]
  N <- pars["N"]; sigma <- pars["sigma"]
  
  t_max <- max(times)
  n_steps <- ceiling(t_max / dt)
  t_all <- seq(0, by = dt, length.out = n_steps + 1)
  
  S <- numeric(n_steps + 1); I <- numeric(n_steps + 1); R <- numeric(n_steps + 1)
  S[1] <- state0["S"]; I[1] <- state0["I"]; R[1] <- state0["R"]
  
  for (k in seq_len(n_steps)) {
    foi <- beta * S[k] * I[k] / N
    rec <- gamma * I[k]
    dW <- rnorm(3) * sqrt(dt)
    
    S[k+1] <- S[k] + (-foi) * dt + sigma * sqrt(abs(foi)) * dW[1]
    I[k+1] <- I[k] + (foi - rec) * dt + sigma * sqrt(abs(foi + rec)) * dW[2]
    R[k+1] <- R[k] + rec * dt + sigma * sqrt(abs(rec)) * dW[3]
  }
  
  # Interpolate to requested times
  idx <- findInterval(times, t_all, all.inside = TRUE)
  data.frame(time = times, S = S[idx], I = I[idx], R = R[idx])
}

sde_pars <- c(beta = 0.5, gamma = 0.1, N = 1000, sigma = 0.1)
out_sde <- sir_sde_euler_maruyama(sde_pars, state0, times, dt = 0.01)

cat("SDE I peak:", max(out_sde$I), "\n")
cat("SDE peak time:", out_sde$time[which.max(out_sde$I)], "\n")
```

## Comparing ODE vs SDE

```{r}
plot(out_ode[, "time"], out_ode[, "I"], type = "l", lwd = 2, col = "black",
     xlab = "Time", ylab = "Infected (I)",
     main = "ODE vs SDE: Infected Compartment")
lines(out_sde$time, out_sde$I, col = "steelblue", lwd = 1.5)
legend("topright", legend = c("ODE (deterministic)", "SDE (Euler-Maruyama)"),
       col = c("black", "steelblue"), lwd = c(2, 1.5))
```

The SDE trajectory fluctuates around the deterministic ODE solution due to the Wiener process noise.

## Ensemble of SDE paths

Running multiple realisations produces an ensemble showing the range of stochastic outcomes:

```{r}
n_particles <- 100
ensemble_I <- matrix(NA, nrow = length(times), ncol = n_particles)

for (p in seq_len(n_particles)) {
  run <- sir_sde_euler_maruyama(sde_pars, state0, times, dt = 0.01, seed = 100 + p)
  ensemble_I[, p] <- run$I
}

I_mean <- rowMeans(ensemble_I)
I_sd <- apply(ensemble_I, 1, sd)

cat("Mean peak I:", max(I_mean), "\n")
cat("Std at peak:", I_sd[which.max(I_mean)], "\n")
```

```{r}
plot(NULL, xlim = range(times), ylim = c(0, max(ensemble_I) * 1.05),
     xlab = "Time", ylab = "Infected (I)", main = "Ensemble of SDE Trajectories")
for (p in seq_len(min(n_particles, 50))) {
  lines(times, ensemble_I[, p], col = adjustcolor("steelblue", alpha.f = 0.15))
}
lines(times, I_mean, col = "red", lwd = 2)
lines(times, I_mean + 2 * I_sd, col = "red", lwd = 1, lty = 2)
lines(times, I_mean - 2 * I_sd, col = "red", lwd = 1, lty = 2)
lines(out_ode[, "time"], out_ode[, "I"], col = "black", lwd = 2)
legend("topright",
       legend = c("ODE", "SDE mean", "SDE ± 2σ", "SDE paths"),
       col = c("black", "red", "red", "steelblue"),
       lty = c(1, 1, 2, 1), lwd = c(2, 2, 1, 1))
```

## SDE: Milstein scheme

The Milstein scheme achieves strong order 1.0 by including the derivative of the diffusion coefficient:

$$
X_{n+1} = X_n + f(X_n)\,\Delta t + \sigma(X_n)\,\Delta W + \frac{1}{2}\sigma(X_n)\sigma'(X_n)\left((\Delta W)^2 - \Delta t\right)
$$

For our SIR model with diffusion $\sigma_X = \sigma\sqrt{|r|}$ where $r$ is the transition rate, the derivative is $\sigma'_X = \sigma / (2\sqrt{|r|})$.

```{r}
sir_sde_milstein <- function(pars, state0, times, dt = 0.01, seed = 42) {
  set.seed(seed)
  beta <- pars["beta"]; gamma <- pars["gamma"]
  N <- pars["N"]; sigma_coef <- pars["sigma"]
  
  t_max <- max(times)
  n_steps <- ceiling(t_max / dt)
  t_all <- seq(0, by = dt, length.out = n_steps + 1)
  
  S <- numeric(n_steps + 1); I <- numeric(n_steps + 1); R <- numeric(n_steps + 1)
  S[1] <- state0["S"]; I[1] <- state0["I"]; R[1] <- state0["R"]
  
  for (k in seq_len(n_steps)) {
    foi <- beta * S[k] * I[k] / N
    rec <- gamma * I[k]
    dW <- rnorm(3) * sqrt(dt)
    
    # Diffusion coefficients
    diff_S <- sigma_coef * sqrt(abs(foi))
    diff_I <- sigma_coef * sqrt(abs(foi + rec))
    diff_R <- sigma_coef * sqrt(abs(rec))
    
    # Milstein correction: 0.5 * sigma * sigma' * (dW^2 - dt)
    # For sigma(r) = sigma_coef * sqrt(|r|), sigma'(r) = sigma_coef / (2*sqrt(|r|))
    # => sigma * sigma' = sigma_coef^2 / 2
    mil_corr <- 0.5 * sigma_coef^2 / 2
    
    S[k+1] <- S[k] + (-foi) * dt + diff_S * dW[1] + mil_corr * (dW[1]^2 - dt)
    I[k+1] <- I[k] + (foi - rec) * dt + diff_I * dW[2] + mil_corr * (dW[2]^2 - dt)
    R[k+1] <- R[k] + rec * dt + diff_R * dW[3] + mil_corr * (dW[3]^2 - dt)
  }
  
  idx <- findInterval(times, t_all, all.inside = TRUE)
  data.frame(time = times, S = S[idx], I = I[idx], R = R[idx])
}

out_mil <- sir_sde_milstein(sde_pars, state0, times, dt = 0.01, seed = 42)
cat("EM final I:       ", tail(out_sde$I, 1), "\n")
cat("Milstein final I: ", tail(out_mil$I, 1), "\n")
```

The Milstein scheme achieves better strong convergence (individual path accuracy) at the same step size, which matters when comparing specific realisations. In Julia, Odin provides both solvers natively via the `:euler_maruyama` and `:milstein` solver options.

## Effect of noise magnitude

The parameter $\sigma$ controls the noise intensity:

```{r}
for (sigma_val in c(0.01, 0.1, 0.5)) {
  p <- c(beta = 0.5, gamma = 0.1, N = 1000, sigma = sigma_val)
  ens <- matrix(NA, nrow = length(times), ncol = 50)
  for (j in 1:50) {
    run <- sir_sde_euler_maruyama(p, state0, times, dt = 0.01, seed = 200 + j)
    ens[, j] <- run$I
  }
  I_m <- rowMeans(ens)
  peak_var <- var(ens[which.max(I_m), ])
  cat(sprintf("σ = %.2f → variance at peak: %.2f\n", sigma_val, peak_var))
}
```

Higher $\sigma$ produces wider spread in trajectories.

## Summary

| Feature | R Implementation | Julia/Odin |
|---------|-----------------|------------|
| ODE (deterministic) | `deSolve::ode()` | `deriv(X) = expr` |
| SDE (Euler-Maruyama) | Manual loop | `diffusion(X) = expr` + `:euler_maruyama` |
| SDE (Milstein) | Manual loop | `diffusion(X) = expr` + `:milstein` |

Key points:

- R requires manual implementation of SDE solvers, while Odin/Julia provides them natively
- `deSolve` handles the ODE baseline efficiently
- The Euler-Maruyama scheme is straightforward to implement (strong order 0.5)
- The Milstein scheme requires computing the derivative of the diffusion coefficient (strong order 1.0)
- Multiple realisations give an ensemble view of stochastic uncertainty
- Julia's Odin advantage: built-in SDE support with `diffusion()` keyword and automatic solver selection
