---
title: "Compositional Model Building"
subtitle: "odin2/dust2 Vignette 07 (R)"
format:
  html:
    toc: true
    code-fold: false
---

## Introduction

This is the R companion to the Julia categorical vignette. While R's odin2 does not have a built-in categorical composition layer, we demonstrate equivalent models built manually, matching the Julia compositional workflow.

## Basic SIR

```{r}
library(odin2)
library(dust2)

sir <- odin({
  deriv(S) <- -beta * S * I / N
  deriv(I) <- beta * S * I / N - gamma * I
  deriv(R) <- gamma * I
  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  beta <- parameter(0.3)
  gamma <- parameter(0.1)
  I0 <- parameter(10)
  N <- parameter(1000)
})

pars <- list(beta = 0.3, gamma = 0.1, I0 = 10, N = 1000)
sys <- dust_system_create(sir, pars, ode_control = dust_ode_control())
dust_system_set_state_initial(sys)
times <- seq(0, 200, by = 0.5)
result <- dust_system_simulate(sys, times)

plot(times, result[1, ], type = "l", col = "blue", lwd = 2,
     xlab = "Time", ylab = "Population", main = "SIR ODE Model",
     ylim = c(0, 1000))
lines(times, result[2, ], col = "red", lwd = 2)
lines(times, result[3, ], col = "green", lwd = 2)
legend("right", legend = c("S", "I", "R"),
       col = c("blue", "red", "green"), lwd = 2)
```

## SIRS — Equivalent to Composed Model

In Julia, this was built by composing infection + recovery + waning immunity sub-models.
In R, we write the full model directly:

```{r}
sirs <- odin({
  deriv(S) <- -beta * S * I / N + delta * R
  deriv(I) <- beta * S * I / N - gamma * I
  deriv(R) <- gamma * I - delta * R
  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  beta <- parameter(0.3)
  gamma <- parameter(0.1)
  delta <- parameter(0.01)
  I0 <- parameter(10)
  N <- parameter(1000)
})

pars <- list(beta = 0.3, gamma = 0.1, delta = 0.01, I0 = 10, N = 1000)
sys <- dust_system_create(sirs, pars, ode_control = dust_ode_control())
dust_system_set_state_initial(sys)
times <- seq(0, 1000, by = 1)
result <- dust_system_simulate(sys, times)

plot(times, result[1, ], type = "l", col = "blue", lwd = 2,
     xlab = "Time", ylab = "Population",
     main = "SIRS — Endemic Equilibrium",
     ylim = c(0, 1000))
lines(times, result[2, ], col = "red", lwd = 2)
lines(times, result[3, ], col = "green", lwd = 2)
legend("right", legend = c("S", "I", "R"),
       col = c("blue", "red", "green"), lwd = 2)
```

## SEIR — Equivalent to Three-Part Composition

```{r}
seir <- odin({
  deriv(S) <- -beta * S * I / N
  deriv(E) <- beta * S * I / N - sigma * E
  deriv(I) <- sigma * E - gamma * I
  deriv(R) <- gamma * I
  initial(S) <- N - I0
  initial(E) <- 0
  initial(I) <- I0
  initial(R) <- 0
  beta <- parameter(0.5)
  sigma <- parameter(0.2)
  gamma <- parameter(0.1)
  I0 <- parameter(10)
  N <- parameter(1000)
})

pars <- list(beta = 0.5, sigma = 0.2, gamma = 0.1, I0 = 10, N = 1000)
sys <- dust_system_create(seir, pars, ode_control = dust_ode_control())
dust_system_set_state_initial(sys)
times <- seq(0, 200, by = 0.5)
result <- dust_system_simulate(sys, times)

plot(times, result[1, ], type = "l", col = "blue", lwd = 2,
     xlab = "Time", ylab = "Population", main = "SEIR Model",
     ylim = c(0, 1000))
lines(times, result[2, ], col = "orange", lwd = 2)
lines(times, result[3, ], col = "red", lwd = 2)
lines(times, result[4, ], col = "green", lwd = 2)
legend("right", legend = c("S", "E", "I", "R"),
       col = c("blue", "orange", "red", "green"), lwd = 2)
```

## Age-Stratified SIR — Equivalent to Stratification

In Julia, we used `stratify(sir, [:child, :adult, :elder]; contact=C)`.
In R, we build the age-structured model with arrays:

```{r}
sir_age <- odin({
  n_age <- parameter(3)
  dim(S) <- n_age
  dim(I) <- n_age
  dim(R) <- n_age
  dim(foi) <- n_age
  dim(S0) <- n_age
  dim(I0) <- n_age
  dim(contact) <- c(n_age, n_age)
  dim(weighted) <- c(n_age, n_age)

  # Force of infection per group: beta * sum_j(C[i,j] * I[j]) / N
  weighted[, ] <- contact[i, j] * I[j]
  foi[] <- beta * sum(weighted[i, ]) / N

  deriv(S[]) <- -foi[i] * S[i]
  deriv(I[]) <- foi[i] * S[i] - gamma * I[i]
  deriv(R[]) <- gamma * I[i]

  initial(S[]) <- S0[i]
  initial(I[]) <- I0[i]
  initial(R[]) <- 0

  S0[] <- parameter()
  I0[] <- parameter()
  contact[, ] <- parameter()
  beta <- parameter(0.15)
  gamma <- parameter(0.1)
  N <- parameter(1000)
})

C <- matrix(c(3.0, 1.0, 0.5,
              1.0, 2.0, 0.5,
              0.5, 0.5, 1.0), 3, 3, byrow = TRUE)

pars <- list(
  n_age = 3,
  S0 = c(325, 325, 330),
  I0 = c(3.33, 3.33, 3.34),
  contact = C,
  beta = 0.15, gamma = 0.1, N = 1000
)

sys <- dust_system_create(sir_age, pars, ode_control = dust_ode_control())
dust_system_set_state_initial(sys)
times <- seq(0, 200, by = 0.5)
result <- dust_system_simulate(sys, times)

cols <- c("blue", "red", "green")
groups <- c("Child", "Adult", "Elder")
plot(NULL, xlim = range(times), ylim = c(0, max(result[4:6, ])),
     xlab = "Time", ylab = "Infected", main = "Infected by Age Group")
for (i in 1:3) {
  lines(times, result[3 + i, ], col = cols[i], lwd = 2)
}
legend("topright", legend = groups, col = cols, lwd = 2)
```

## SIR + Vaccination

```{r}
sir_vax <- odin({
  deriv(S) <- -beta * S * I / N - nu * S
  deriv(I) <- beta * S * I / N - gamma * I
  deriv(R) <- gamma * I
  deriv(V) <- nu * S
  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(V) <- 0
  beta <- parameter(0.3)
  gamma <- parameter(0.1)
  nu <- parameter(0)
  I0 <- parameter(10)
  N <- parameter(1000)
})

times <- seq(0, 200, by = 0.5)

# Without vaccination
sys1 <- dust_system_create(sir_vax, list(beta = 0.3, gamma = 0.1, nu = 0,
                                          I0 = 10, N = 1000),
                           ode_control = dust_ode_control())
dust_system_set_state_initial(sys1)
r1 <- dust_system_simulate(sys1, times)

# With vaccination
sys2 <- dust_system_create(sir_vax, list(beta = 0.3, gamma = 0.1, nu = 0.005,
                                          I0 = 10, N = 1000),
                           ode_control = dust_ode_control())
dust_system_set_state_initial(sys2)
r2 <- dust_system_simulate(sys2, times)

plot(times, r1[2, ], type = "l", col = "red", lwd = 2,
     xlab = "Time", ylab = "Infected",
     main = "Effect of Vaccination",
     ylim = c(0, max(r1[2, ])))
lines(times, r2[2, ], col = "blue", lwd = 2)
legend("topright", legend = c("No vaccination", "ν = 0.005/day"),
       col = c("red", "blue"), lwd = 2)
```

## Summary

| Feature | Julia (Odin.jl) | R (odin2) |
|---------|-----------------|-----------|
| Model definition | `EpiNet` Petri nets | Manual DSL |
| Composition | `compose(net1, net2)` | Manual |
| Stratification | `stratify(net, groups; contact=C)` | Arrays + loops |
| Lowering | `lower(net; mode=:ode)` | N/A (already in DSL) |

The Julia categorical extension automates what must be done manually in R, reducing errors and enabling rapid prototyping of complex model structures.
